\section{Background}
\label{sec:background}

This section presents the theoretical foundations that ground the
taxonomy developed in Section~\ref{sec:taxonomy}. We draw from three
domains: cognitive neuroscience, information retrieval, and Yogac\=ara
Buddhist psychology.

\subsection{Cognitive Neuroscience of Memory}

\subsubsection{Complementary Learning Systems}
McClelland~\etal~\cite{mcclelland1995} proposed that the brain uses two
complementary systems to avoid catastrophic forgetting: a fast learner
(hippocampus) that records raw episodes and a slow learner (neocortex)
that extracts stable patterns through interleaved replay.

\subsubsection{Hebbian Learning}
``Neurons that fire together wire together''~\cite{hebb1949}. Long-term
potentiation (LTP) and long-term depression (LTD) provide the
biological basis for associative memory~\cite{bliss1973}.

\subsubsection{Spreading Activation}
Collins and Loftus~\cite{collins1975} showed that activating a concept
propagates through a weighted semantic network, priming associated
concepts. This explains how ``doctor'' activates ``nurse'' even without
direct keyword overlap.

\subsubsection{Bjork Dual-Strength Forgetting}
Each memory has two independent strengths: \emph{storage strength} (how
well-encoded) and \emph{retrieval strength} (how accessible
now)~\cite{bjork1992}. Retrieval strength decays over time; storage
strength only increases with access. This model explains tip-of-the-tongue
phenomena and why relearning is faster than initial
learning. Anderson~\etal~\cite{anderson1994} showed that retrieving a
memory suppresses competing memories sharing the same cues
(retrieval-induced forgetting, RIF).

\subsection{Information Retrieval}

\subsubsection{Reciprocal Rank Fusion}
RRF merges multiple ranked result sets without score
calibration~\cite{cormack2009rrf}: $\text{score}(d) = \sum_i 1 / (k +
\text{rank}_i + 1)$. Bruch~\etal~\cite{bruch2023} showed that convex
combination outperforms RRF when tuning data is available, but RRF is
superior in zero-shot settings---the typical case for agent memory.

\subsubsection{Encoding Specificity}
Tulving and Thomson~\cite{tulving1973} demonstrated that a memory is
most accessible when the retrieval context matches the encoding
context. A word encoded in the presence of a weak associate (e.g.,
``train'' encoded with ``black'') is better recalled when that same
associate is provided at retrieval, even if a strong associate
(``white'') is substituted instead. This \emph{encoding specificity
principle} implies that memory systems should preserve the context in
which information was originally stored---topics, participants,
sentiment, temporal position---and exploit that context at retrieval
time. The principle bridges neuroscience and IR: it motivates context-aware
reranking and the multi-signal fusion architectures found in several of
the systems surveyed in Section~\ref{sec:systems}.

\subsection{Yogac\=ara Buddhist Psychology}

The Yogac\=ara school of Buddhist philosophy, systematized in the
\emph{Yogac\=arabh\=umi-\\'s\=astra} attributed to
Maitreya/Asa\.{n}ga and the \emph{Cheng Weishi Lun} (\cjk{成唯識論})
by Xuanzang (659~CE), proposes eight consciousnesses. The eighth,
\emph{\=alaya-vij\~n\=ana} (\cjk{阿賴耶識}, storehouse consciousness),
is a persistent substrate holding all \emph{b\=ija} (seeds)---latent
potentials that ripen into experience when conditions
converge~\cite{yogacarabhumi, chengweishilun}. The framework provides a
vocabulary for phenomena that cognitive neuroscience addresses piecemeal:
persistent latent storage, gradual trace accumulation, and periodic
structural reorganization.

\subsubsection{The Storehouse and Seeds}

The \emph{\=alaya-vij\~n\=ana} (\cjk{阿賴耶識}) is the eighth and
deepest consciousness in Yogac\=ara theory---a continuous stream that
underlies all mental activity, persisting through sleep, dreamless
states, and (in the traditional account) between lifetimes. It receives
impressions from every experience and stores them as
\emph{b\=ija} (\cjk{種子}, seeds): latent potentials that are not
static data but living dispositions with conditions for ripening.

The key verse from the \emph{Sa\d{m}dhinirmocana S\=utra}
(\cjk{解深密經}) captures this character~\cite{samdhinirmocana}:

\begin{quote}
\cjk{阿陀那識甚深細，一切種子如瀑流，我於凡愚不開演，恐彼分別執為我}

\medskip
\emph{The \=ad\=ana consciousness is exceedingly deep and subtle; all
its seeds are like a torrential flood. I do not reveal it to the
foolish, for fear they would grasp it as a self.}
\end{quote}

The metaphor of a torrential flood emphasizes that the storehouse is not
a passive archive but a dynamic flow. Seeds interact with one another:
they have mutual dependence (\emph{sahabhuh\=etu}), they ripen
heterogeneously (\emph{vip\=aka}, \cjk{異熟}---the result differs in
nature from the cause), and they are continuously modified by new
experience. A seed ripens into manifest experience when appropriate
conditions converge, much as retrieval strength in the Bjork model
determines whether a latent memory surfaces.

For memory system design, this framework suggests two principles:
(i)~stored items should be treated as potentials with activation
conditions rather than inert records, and (ii)~a memory system benefits
from maintaining the distinction between the \emph{persistence} of
storage (seeds remain) and the \emph{accessibility} of retrieval (seeds
ripen only when conditions align).

\subsubsection{Perfuming and Transformation}

\emph{V\=asan\=a} (\cjk{薰習}, perfuming) is the process by which each
experience leaves a subtle trace on the storehouse, like incense
gradually permeating cloth~\cite{chengweishilun}. No single experience
is transformative on its own; rather, the accumulation of many similar
traces shapes disposition and behavior. The \emph{Cheng Weishi Lun}
distinguishes \emph{ming-yan v\=asan\=a} (name-and-word perfuming,
\cjk{名言薰習})---linguistic and conceptual traces---from
\emph{ye v\=asan\=a} (action perfuming, \cjk{業薰習}), which records
the volitional quality of action. Both accumulate incrementally, and
both shape future experience.

The complementary process is \emph{\=a\'sraya-par\=av\d{r}tti}
(\cjk{轉依}, transformation of the basis): a periodic restructuring of
the storehouse itself. Where perfuming is incremental addition,
transformation is qualitative reorganization---resolving contradictions,
pruning stale dispositions, and moving the substrate toward greater
coherence. In the Yogac\=ara soteriological context, the ultimate
transformation converts the eight consciousnesses into four wisdoms
(\cjk{轉識成智}): the storehouse becomes \emph{great mirror wisdom}
(\cjk{大圓鏡智}), reflecting reality without distortion.

These two operations---incremental perfuming and periodic
transformation---map directly onto patterns observed across
agent memory systems. Perfuming corresponds to trace accumulation:
strengthening associations on co-retrieval (Hebbian LTP), recording
implicit preferences from behavioral patterns, and updating retrieval
strength on access (Bjork). Transformation corresponds to consolidation
and maintenance: deduplicating near-identical entries, decaying
unreinforced associations (Hebbian LTD), pruning stale records, and
restructuring knowledge graphs. The distinction between the two
operations provides a useful analytical frame for the taxonomy
dimensions developed in Section~\ref{sec:taxonomy}.
