\section{System-by-System Analysis}
\label{sec:systems}

We organize systems into six categories: production systems
(\S\ref{sec:production}), standalone memory servers
(\S\ref{sec:standalone}), framework-level memory
(\S\ref{sec:framework}), coding agent memory (\S\ref{sec:coding-agents}),
file-based memory (\S\ref{sec:filebased}), and research architectures
(\S\ref{sec:research}). For each system we identify the memory
types supported (mapped to the CoALA taxonomy~\cite{sumers2024coala}),
the retrieval architecture (mapped to the RAG
taxonomy~\cite{gao2023rag}), the lifecycle operations, and the
infrastructure requirements. Quantitative cross-comparisons appear in
\S\ref{sec:comparison}; here we focus on what makes each system
architecturally distinct.

%% ===============================================================
\subsection{Production Systems}
\label{sec:production}

\paragraph{Mem0.}
Mem0~\cite{choudhary2025mem0} implements a tiered memory architecture
backed by vector databases (Qdrant, Pinecone, or Chroma), a relational
store for metadata, and an optional Neo4j graph layer (the Mem0g
variant). Memory formation is LLM-dependent: every incoming
interaction is routed through an LLM that extracts facts, user
preferences, and personality traits, then decides whether to create,
update, or deduplicate entries. Retrieval combines vector similarity
with graph traversal and contextual priority scoring, placing Mem0 in
the Advanced RAG category. Forgetting follows a single-curve
exponential decay of low-relevance entries, with periodic consolidation
between short-term and long-term tiers. The authors report a 26\%
accuracy improvement over OpenAI's built-in memory on the LOCOMO
benchmark and 91\% lower p95 latency versus full-context approaches.

The key architectural tradeoff in Mem0 is its heavy infrastructure
requirement (2--3 external services) and its reliance on an LLM for
every memory write. This makes Mem0 well-suited for cloud-deployed,
multi-user SaaS products with dedicated operations teams, but less
appropriate for privacy-first or offline use cases. The Mem0g graph
variant adds entity-relationship traversal, but the graph is
static---entities and relationships exist only when the LLM explicitly
creates them, with no dynamic restructuring through use.

\paragraph{Zep / Graphiti.}
Graphiti~\cite{dempsey2025graphiti}, the engine underlying Zep's memory
service, constructs a hierarchical temporal knowledge graph on Neo4j
with three subgraph layers: community (cluster summaries), entity
(semantic entities and relationships), and episodic (raw conversation
episodes). Retrieval uses a triple-hybrid pipeline---cosine similarity,
Okapi BM25, and breadth-first graph traversal---fused via Reciprocal
Rank Fusion (RRF) with subsequent maximal marginal relevance (MMR)
diversification. Multi-stage reranking considers episode mentions, node
distance, and cross-signal fusion. This places Zep/Graphiti firmly in
the Modular RAG category, alongside a small minority of systems
(18.2\%, Table~\ref{tab:rag-distribution}).

Zep's most distinctive lifecycle feature is \emph{bi-temporal
invalidation}: facts are never deleted, only marked with validity
periods recording when each fact was true in the real world and when
the system learned it. This conservative approach preserves full
provenance but means the knowledge graph grows monotonically. Entity
extraction requires an LLM (typically \texttt{gpt-4o-mini}), though
retrieval itself is LLM-free. The authors report 94.8\% accuracy on
the DMR benchmark and up to 18.5\% accuracy improvement on
LongMemEval.

\paragraph{Letta (formerly MemGPT).}
Letta~\cite{packer2023memgpt} introduced the operating-system metaphor
for agent memory: core memory serves as ``RAM'' (always in-context,
editable by the agent), recall memory stores conversation history, and
archival memory provides unbounded ``disk'' storage via a vector or
graph database. The defining architectural choice is that the LLM
\emph{is} the memory manager---it decides what to store, evict, and
retrieve via explicit tool calls (\texttt{core\_memory\_append},
\texttt{archival\_memory\_search}, etc.).

This agent-driven design makes Letta's memory quality entirely
dependent on LLM capability. When the context window fills, the
system evicts approximately 70\% of messages through summarization.
A 2025 extension adds ``sleep-time agents'' that handle asynchronous
reorganization of archival memory outside the main interaction loop.
Letta's CoALA mapping spans all three memory types (working, episodic,
semantic), but its retrieval is Naive RAG---agent-driven tool calls
with single-signal vector similarity. The OS metaphor proved
influential: MemoryOS~\cite{kang2025memoryos} and OpenViking both
adopted tiered hierarchies, though with different management
strategies.

\paragraph{Hindsight.}
Hindsight~\cite{latimer2025hindsight} introduces the most novel memory
decomposition in the production landscape: four logical networks for
\emph{world} (objective facts), \emph{experience} (agent
interactions), \emph{opinion} (subjective beliefs with confidence
scores), and \emph{observation} (preference-neutral entity summaries).
The opinion network is architecturally unique---each belief carries an
explicit confidence score that evolves as new evidence arrives,
enabling the system to distinguish between well-established facts and
tentative assessments.

Retrieval uses two specialized components: Tempr (temporal entity
memory priming) for time-aware recall and Cara (coherent adaptive
reasoning) for multi-step inference across networks. Hindsight does
not implement explicit forgetting; instead, belief confidence scores
decay or strengthen as evidence accumulates, producing an emergent
form of selective retention. The authors claim state-of-the-art
results: 91.4\% on LongMemEval and 89.61\% on LoCoMo. The four-network
decomposition is finer-grained than most systems' two- or three-store
architectures, but Hindsight lacks a graph overlay and does not
implement principled forgetting beyond confidence evolution.

%% ===============================================================
\subsection{Standalone Memory Servers}
\label{sec:standalone}

Several standalone servers provide memory as a network service,
typically via REST or MCP endpoints, without being tied to a specific
agent framework. \textbf{Motorhead} (Metal, Rust) implements a
minimal conversation buffer with incremental summarization on Redis:
when a configurable window (default 12 messages) fills, the oldest
half is summarized and the summary is incrementally updated.
\textbf{Engram} (Go) takes an agent-trusting approach---the LLM
client decides what to remember, storing curated summaries in SQLite
with FTS5 full-text search and no vector retrieval. Both represent
Naive RAG with no graph, no forgetting model, and no consolidation.
\textbf{OpenViking} (ByteDance/Volcengine, open-sourced January 2026)
adopts a virtual filesystem metaphor, mapping all agent context to a
\texttt{viking://} protocol with three tiers: L0 (immediate context),
L1 (session-level), and L2 (persistent archival). Built on VikingDB
infrastructure, OpenViking replaces fragmented vector storage with
hierarchical directory-based retrieval, but provides no forgetting,
no graph dynamics, and no preference learning.

%% ===============================================================
\subsection{Framework-Level Memory}
\label{sec:framework}

The major agent frameworks provide memory as a built-in module, but
these integrations are typically shallow---memory as an afterthought
rather than a first-class architectural concern. \textbf{LangChain}'s
memory classes (ConversationBufferMemory, ConversationSummaryMemory,
ConversationBufferWindowMemory) offer working-memory-only storage with
truncation-based forgetting; most are now deprecated in favor of
\texttt{RunnableWithMessageHistory}. \textbf{LlamaIndex} provides
composable memory blocks (StaticMemoryBlock,
FactExtractionMemoryBlock, VectorMemoryBlock) with FIFO eviction and
a configurable token ratio (70\% chat history, 30\% long-term by
default). \textbf{LangMem SDK}, LangChain's dedicated memory engine,
adds semantic memory (extracted facts and preferences) and a
distinctive \emph{procedural} memory capability: learned patterns are
saved as updated prompt instructions, modifying the agent's behavior
without explicit retrieval. None of these frameworks implements a
graph overlay, principled forgetting, consolidation, or preference
emergence. Their retrieval ranges from direct prompt injection
(LangChain) to vector similarity via LangGraph's storage layer
(LangMem), placing them uniformly in the Naive RAG category.

%% ===============================================================
\subsection{Coding Agent Memory}
\label{sec:coding-agents}

A distinctive cluster of memory systems has emerged around coding
agents (Claude Code, Cursor, Windsurf), exposed primarily via the
Model Context Protocol (MCP). These systems share a common pattern:
replace the default full-context injection of MEMORY.md files with
ranked retrieval, typically achieving 70--96\% token savings.
\textbf{memsearch}~\cite{zilliz2026a} (Zilliz/Milvus team) extracts
the default memory file structure and overlays BM25 plus vector
retrieval with RRF fusion and progressive top-$k$ disclosure.
\textbf{QMD}~\cite{lutke2026qmd} (Tobi L\"utke, Shopify) runs three
local GGUF models for embedding (Gemma-300M), reranking
(Qwen3-Reranker-0.6B), and query expansion (1.7B parameter model),
achieving fully offline hybrid retrieval with cross-encoder reranking
and reported 96\% token reduction~\cite{levine2026}. \textbf{Beads}
(Go, on Dolt version-controlled SQL) models a dependency-aware issue
graph with four link types, enabling traversal along dependency chains
rather than pure similarity. \textbf{Basic Memory} (Python) takes a
wiki-link approach, storing structured Markdown with explicit
Observations and Relations that form a semantic knowledge graph
navigable via \texttt{memory://} URLs. All four independently adopted
RRF over the default weighted linear fusion---a convergence we
analyze further in \Cref{sec:memorymd}.

%% ===============================================================
\subsection{File-Based Memory}
\label{sec:filebased}

The simplest memory architecture stores knowledge in Markdown files
that the agent reads and writes directly. The dominant instance is the
\textbf{MEMORY.md pattern} used by coding agents such as OpenClaw:
a curated long-term knowledge file (MEMORY.md) plus daily interaction
logs with temporal decay, backed by a SQLite index with FTS5 and
sqlite-vec for hybrid retrieval (70\% vector, 30\% BM25 weighted
fusion). \textbf{Claudesidian} extends this pattern into Obsidian
vaults, using PARA folder organization and daily/weekly notes with
Obsidian's wiki-link graph for navigation. File-based memory provides
zero-infrastructure deployment and human-readable persistence, but
scales poorly: MEMORY.md is injected into the system prompt in full
on every turn regardless of relevance, consuming approximately 35,600
tokens per message (93.5\% waste in multi-message
sessions)~\cite{openclaw9157}. We analyze the cost, structure, and
retrieval failure modes of file-based memory in detail in
\Cref{sec:memorymd}.

%% ===============================================================
\subsection{Research Architectures}
\label{sec:research}

\paragraph{Generative Agents.}
Park et al.~\cite{park2023generative} introduced the seminal memory
architecture for LLM agents: a chronological \emph{memory stream} of
observations, from which the agent periodically generates
\emph{reflections}---higher-order thoughts that synthesize multiple
observations into abstract insights. Retrieval scores each memory by a
weighted product of three signals: \emph{recency} (exponential decay),
\emph{importance} (LLM-rated 1--10), and \emph{relevance} (cosine
similarity to the current query). This triple-scored formula became the
foundational retrieval mechanism for agent memory, adopted or adapted
by the majority of subsequent systems.

The architecture's significance lies less in its individual components
(all relatively simple) than in their composition: the reflection
mechanism implements a form of episodic-to-semantic consolidation,
where accumulated observations are distilled into reusable knowledge.
The system uses in-memory storage (ephemeral), a flat memory stream
(no typed stores), no graph overlay, and single-curve recency
decay---limitations that subsequent work has systematically addressed.
The reflection mechanism, however, inspired consolidation pipelines
in multiple production and research systems.

\paragraph{HippoRAG.}
HippoRAG~\cite{gutierrez2024hipporag} grounds its architecture in
hippocampal indexing theory from neuroscience: the LLM acts as
neocortex (processing and generating language), a parahippocampal
region (PHR) encoder detects synonymy and entity co-reference, and an
open knowledge graph acts as hippocampus (indexing and associating
memories). Retrieval uses Personalized PageRank (PPR) on the knowledge
graph, propagating activation from query-matched nodes to
transitively connected entities. This enables multi-hop reasoning that
pure embedding similarity cannot achieve---asking ``who handles auth
permissions?'' can retrieve information about ``Alice manages the auth
team'' through entity-relationship traversal rather than lexical
overlap.

HippoRAG achieves 20\% improvement on multi-hop question answering at
10--30$\times$ lower cost than iterative retrieval
approaches. HippoRAG~2~\cite{gutierrez2025rag2memory} extends this
with online knowledge graph updates and a deeper integration of the
complementary learning systems (CLS) framework. The knowledge graph
is static (constructed by LLM extraction), and the system has no
forgetting, consolidation, or preference learning---it is a retrieval
architecture rather than a full memory lifecycle system. PPR on a
knowledge graph is comparable in cognitive inspiration to spreading
activation on associative graphs, though the two mechanisms differ:
PPR computes a global stationary distribution, while spreading
activation models local, time-bounded propagation~\cite{collins1975}.

\paragraph{SYNAPSE.}
SYNAPSE~\cite{jiang2026synapse} constructs a unified episodic-semantic
graph and retrieves through two biologically inspired mechanisms:
\emph{spreading activation} (activation propagates from query-matched
nodes to neighbors through weighted edges, priming associated
memories) and \emph{lateral inhibition} (strongly activated nodes
suppress weakly activated competitors, sharpening retrieval focus).
Retrieval fuses geometric embedding similarity with activation-based
graph traversal in a triple-hybrid pipeline. Temporal decay weakens
links over time.

SYNAPSE achieves +7.2 F1 over prior state of the art on LoCoMo, 23\%
improvement on multi-hop reasoning, and 95\% token reduction versus
full-context approaches. The lateral inhibition mechanism is
particularly notable: it implements a form of retrieval-induced
forgetting~\cite{anderson1994}, where retrieving one memory
suppresses competing memories sharing the same cues. This is the most
neurobiologically grounded retrieval mechanism in the surveyed
systems, though SYNAPSE does not implement consolidation, principled
forgetting (beyond temporal decay), or preference learning. The
system is in-memory Python research code, not a persistent or
embeddable library.

\paragraph{Mem-alpha.}
Mem-alpha~\cite{wang2025memalpha} takes a fundamentally different
approach to memory management: instead of hand-crafting lifecycle
heuristics, it trains memory policies via reinforcement learning.
The architecture decomposes memory into three components---core
(always in context), episodic (interaction history), and semantic
(distilled knowledge)---a decomposition nearly identical to the CoALA
framework. An RL agent learns when to store, what to consolidate,
and what to forget, using downstream question-answering accuracy as
the reward signal. Trained on 30K tokens, the learned policy
generalizes to 400K+ tokens (13$\times$ extrapolation).

The RL-learned approach represents the most significant methodological
divergence from hand-crafted cognitive models. Learned policies can
potentially discover non-obvious management strategies that human
designers would not specify, but they require training data, are
opaque (the policy network's decisions are not interpretable), and may
not transfer across domains. Mem-alpha is part of a broader 2025--2026
trend: Memory-R1, MemRL~\cite{zhang2026memrl}, and AgeMem all train
memory management via RL, collectively establishing learned memory
policies as a distinct paradigm alongside rule-based and
neuroscience-inspired approaches.
