\section{Comparative Analysis}
\label{sec:comparison}

This section provides a quantitative cross-sectional analysis of the 88 memory
systems surveyed in \S\ref{sec:systems}, organized along three axes: feature
prevalence across the full landscape, temporal trends showing how the field has
evolved from 2023 to 2026, and the distribution of retrieval sophistication
mapped to the RAG taxonomy of Gao et al.~\cite{gao2023rag}. We exclude vector
databases (Pinecone, ChromaDB, Weaviate, Milvus, etc.) from all counts, as
these are infrastructure components rather than memory architectures. The
remaining 88 systems span five categories: Dedicated Memory Systems (27),
Framework Memory Modules (13), Coding Agent Memory (12), File-Based Memory (2),
and Research Architectures (12), plus 22 additional systems described in
sufficient detail to classify but not included in the primary comparison
tables.

%% ---------------------------------------------------------------
\subsection{Feature Prevalence}
\label{sec:feature-prevalence}

Table~\ref{tab:feature-prevalence} reports the count and percentage of systems
exhibiting each capability across the full survey. Features are organized into
six groups corresponding to the analytical framework introduced in
\S\ref{sec:background}: memory types drawn from the CoALA taxonomy
\cite{sumers2024coala}, retrieval signals mapped to the RAG survey
\cite{gao2023rag}, lifecycle operations from the memory operations taxonomy
\cite{zhang2024survey}, preference learning, graph structure, and
infrastructure requirements.

\begin{table}[!t]
\centering
\caption{Feature prevalence across 88 surveyed memory systems (excluding vector databases). Counts reflect systems with explicit support for each capability based on documentation, publications, and source code analysis.}
\label{tab:feature-prevalence}
\small
\begin{tabular}{@{}lrr@{}}
\toprule
\textbf{Feature} & \textbf{Count} & \textbf{\%} \\
\midrule
\multicolumn{3}{@{}l}{\textit{Memory Types (CoALA)}} \\
\quad Has episodic memory          & 48 & 54.5 \\
\quad Has semantic memory          & 45 & 51.1 \\
\quad Has procedural memory        &  6 &  6.8 \\
\quad Has implicit/preference mem. & 27 & 30.7 \\
\quad Has working memory           & 31 & 35.2 \\
\midrule
\multicolumn{3}{@{}l}{\textit{Retrieval Signals}} \\
\quad Uses BM25/sparse search      & 19 & 21.6 \\
\quad Uses vector/dense search     & 59 & 67.0 \\
\quad Uses graph traversal         & 28 & 31.8 \\
\quad Uses hybrid fusion (RRF or similar) & 14 & 15.9 \\
\quad Uses reranking               & 12 & 13.6 \\
\midrule
\multicolumn{3}{@{}l}{\textit{Lifecycle Operations}} \\
\quad Has any forgetting mechanism & 34 & 38.6 \\
\quad Has consolidation (ep.\ $\to$ sem.) & 16 & 18.2 \\
\quad Has contradiction resolution &  7 &  8.0 \\
\quad Has transformation/reorganization & 11 & 12.5 \\
\midrule
\multicolumn{3}{@{}l}{\textit{Preference Learning}} \\
\quad Has any preference learning  & 27 & 30.7 \\
\quad \quad via LLM extraction     & 16 & 18.2 \\
\quad \quad via emergent/accumulated & 4 &  4.5 \\
\midrule
\multicolumn{3}{@{}l}{\textit{Graph Structure}} \\
\quad Has any graph structure      & 28 & 31.8 \\
\quad \quad Static KG (LLM-built)  & 20 & 22.7 \\
\quad \quad Dynamic/Hebbian (usage-adaptive) & 4 & 4.5 \\
\midrule
\multicolumn{3}{@{}l}{\textit{Infrastructure Requirements}} \\
\quad Requires 0 external services & 28 & 31.8 \\
\quad Requires 1--2 external services & 37 & 42.0 \\
\quad Requires 3+ external services &  6 &  6.8 \\
\quad Requires LLM for memory ops  & 62 & 70.5 \\
\bottomrule
\end{tabular}
\end{table}

Three patterns emerge from Table~\ref{tab:feature-prevalence}. First, episodic
and semantic memory are roughly equally prevalent (54.5\% and 51.1\%), yet only
18.2\% of systems implement consolidation---the process by which episodic
experiences are distilled into semantic knowledge. This gap is significant:
without consolidation, episodic memories accumulate indefinitely, degrading
retrieval precision as the memory store grows. The few systems that implement
consolidation---Generative Agents' reflection generation
\cite{park2023generative}, LightMem's sleep-time processing
\cite{fang2025lightmem}, SYNAPSE's activation-based promotion
\cite{jiang2026synapse}, and CLS-inspired pipelines---draw on complementary
learning systems theory from neuroscience \cite{mcclelland1995}, suggesting
that biologically grounded approaches to consolidation are more robust than
ad-hoc summarization.

Second, the retrieval landscape shows a pronounced asymmetry: 67.0\% of systems
use vector/dense search, but only 21.6\% combine it with sparse (BM25) search,
and a mere 15.9\% employ principled fusion (typically Reciprocal Rank Fusion).
Full three-signal retrieval---sparse, dense, and graph---with fusion is
implemented by only four systems. This is notable because hybrid retrieval
consistently outperforms single-signal approaches in zero-shot settings where
labeled relevance judgments are unavailable, which is the typical case for agent
memory.

Third, lifecycle operations remain the most underserved area. Forgetting is
present in only 38.6\% of systems, and of those, the majority use simple
mechanisms (window truncation, FIFO eviction, or single-curve exponential
decay). Contradiction resolution (8.0\%) and transformation/reorganization
(12.5\%) are rarer still. The field has invested heavily in memory
\emph{formation} and \emph{retrieval} while largely neglecting memory
\emph{management}---the processes that keep a memory store healthy as it grows.

%% ---------------------------------------------------------------
\subsection{Temporal Trends}
\label{sec:temporal-trends}

Table~\ref{tab:temporal-trends} groups the surveyed systems by year of initial
publication or release and tracks the adoption of four capabilities that
distinguish sophisticated memory architectures from simple retrieval stores:
graph structure, forgetting, consolidation, and zero-dependency deployment.

\begin{table}[!t]
\centering
\caption{Temporal trends in memory system capabilities by year of publication or initial release. Percentages are relative to the total systems released in that year.}
\label{tab:temporal-trends}
\small
\begin{tabular}{@{}lrrrrrr@{}}
\toprule
& & \multicolumn{4}{c}{\textbf{\% of systems with capability}} \\
\cmidrule(l){3-6}
\textbf{Year} & \textbf{$N$} & \textbf{Graph} & \textbf{Forgetting} & \textbf{Consolidation} & \textbf{0-dep} \\
\midrule
2023 &  9 & 11.1 & 33.3 & 22.2 & 33.3 \\
2024 & 14 & 28.6 & 28.6 & 14.3 & 28.6 \\
2025 & 42 & 33.3 & 40.5 & 19.0 & 31.0 \\
2026 & 23 & 39.1 & 47.8 & 21.7 & 30.4 \\
\bottomrule
\end{tabular}
\vspace{0.3em}
\begin{flushleft}
\footnotesize
\textit{Note:} ``0-dep'' = requires zero external services (no database servers, no Docker, no cloud APIs). Year is determined by first public release or arXiv preprint date.
\end{flushleft}
\end{table}

Table~\ref{tab:temporal-trends} reveals two countervailing trends. Graph
structure has seen steady adoption, rising from 11.1\% in 2023 to 39.1\% in
2026---a trajectory consistent with the ``graph memory is table stakes''
consensus that emerged in early 2026. Similarly, forgetting mechanisms have
grown from 33.3\% to 47.8\%, driven both by Ebbinghaus-curve implementations in
production systems and by RL-learned decay policies in academic work.

In contrast, consolidation (episodic $\to$ semantic promotion) and
zero-dependency deployment have remained effectively flat across all four
years. Consolidation hovers near 20\%, suggesting that this capability
requires deep architectural commitment that most systems---especially
frameworks and coding agent tools---do not prioritize. Zero-dependency
deployment has actually declined slightly in relative terms, as the
2025--2026 wave of production systems (Mem0, Supermemory, TiMem, Papr
Memory) trend toward multi-service architectures with Neo4j, Redis, and
managed vector databases.

The 2025--2026 period also marks the emergence of RL-trained memory
policies as a distinct paradigm. Memory-R1, MemRL, AgeMem, and
Mem-alpha~\cite{wang2025memalpha} all appeared in this window, collectively
demonstrating that reinforcement learning can discover effective memory
management strategies without hand-crafted heuristics. This represents the
most significant methodological shift in the field since the seminal
Generative Agents work \cite{park2023generative} established the
recency-importance-relevance retrieval formula in 2023.

%% ---------------------------------------------------------------

Table~\ref{tab:rag-distribution} maps the surveyed systems to the three-tier
RAG taxonomy of Gao et al.~\cite{gao2023rag}: Naive RAG (single retrieval
signal, no fusion), Advanced RAG (multiple signals or sophisticated
scoring, but fixed pipeline), and Modular RAG (composable retrieval
components with principled fusion and post-retrieval processing).

\begin{table}[!t]
\centering
\caption{Distribution of systems across the RAG taxonomy \cite{gao2023rag}. Classification is based on the retrieval pipeline as documented; systems that delegate retrieval entirely to an LLM (e.g., Letta's agent-driven tool calls) are classified as Naive with an agent-augmented qualifier.}
\label{tab:rag-distribution}
\small
\begin{tabular}{@{}lrrp{4.8cm}@{}}
\toprule
\textbf{RAG Category} & \textbf{Count} & \textbf{\%} & \textbf{Representative Systems} \\
\midrule
Naive RAG       & 42 & 47.7 & LangChain, LlamaIndex, Motorhead, Engram, Agency Swarm, MemoryMesh \\
Advanced RAG    & 30 & 34.1 & Mem0, Generative Agents, Hindsight, Memvid, A-MEM, CrewAI \\
Modular RAG     & 16 & 18.2 & Zep/Graphiti, SYNAPSE, HippoRAG, PageIndex, Vestige, QMD \\
\bottomrule
\end{tabular}
\end{table}

Nearly half (47.7\%) of memory systems use Naive RAG---typically a single
vector similarity query or direct context injection with no fusion or
reranking. This is especially common in framework memory modules (LangChain,
LlamaIndex, Haystack, Agency Swarm) and coding agent tools (Engram,
MemoryMesh, ClaudeHistory Cloud), where memory is a secondary concern appended
to a larger orchestration framework. Advanced RAG systems (34.1\%) add
multi-signal scoring (e.g., Generative Agents' recency $\times$ importance
$\times$ relevance formula) or sophisticated single-signal pipelines (e.g.,
Memvid's Tantivy FTS + HNSW with chronological reranking). Modular RAG
(18.2\%) requires composable, independently configurable retrieval components
with principled fusion---a design pattern found almost exclusively in dedicated
memory systems and research architectures that treat retrieval as a first-class
architectural concern.

%% ---------------------------------------------------------------
\subsection{Benchmark Cross-Comparison}
\label{sec:benchmark-cross-comparison}

\input{sections/benchmark-table-fragment}

%% ---------------------------------------------------------------
\subsection{The Long Tail Problem}
\label{sec:long-tail}

The feature prevalence data in Table~\ref{tab:feature-prevalence} reveals a
pronounced long-tail distribution: most capabilities are concentrated in a
small number of systems, while the majority of the 88 surveyed systems
implement only basic memory functionality. To quantify this, we define a
\emph{capability score} as the count of features present from a set of eight
discriminating capabilities: episodic memory, semantic memory, graph structure,
forgetting, consolidation, contradiction resolution, preference learning, and
hybrid retrieval fusion. Table~\ref{tab:capability-distribution} shows the
distribution.

\begin{table}[!t]
\centering
\caption{Distribution of capability scores across 88 systems. The capability score counts the presence of eight discriminating features: episodic memory, semantic memory, graph structure, forgetting, consolidation, contradiction resolution, preference learning, and hybrid retrieval fusion.}
\label{tab:capability-distribution}
\small
\begin{tabular}{@{}crrl@{}}
\toprule
\textbf{Score} & \textbf{Systems} & \textbf{\%} & \textbf{Examples} \\
\midrule
0--1 & 24 & 27.3 & Agency Swarm, Qwen-Agent, AutoGPT \\
2--3 & 33 & 37.5 & LangChain, Engram, Memvid, HippoRAG \\
4--5 & 20 & 22.7 & Mem0, Hindsight, SYNAPSE, LightMem \\
6--7 &  9 & 10.2 & Zep/Graphiti, EverMemOS, MemOS, Vestige \\
8    &  2 &  2.3 & --- \\
\bottomrule
\end{tabular}
\end{table}

The distribution is sharply skewed: 64.8\% of systems score 3 or below,
meaning they implement at most three of the eight capabilities. Only 12.5\% of
systems score 6 or above. This concentration has three implications for the
field.

First, the median memory system is a \textbf{vector store with optional
context windowing}---it stores text, retrieves by embedding similarity, and
manages capacity through truncation or FIFO eviction. The CoALA framework
\cite{sumers2024coala} classifies this as working memory with optional
episodic storage, the simplest configuration in its taxonomy. The gap between
what the literature proposes (multi-store architectures with lifecycle
processes) and what practitioners ship (single-store vector retrieval) remains
substantial.

Second, capabilities cluster rather than distribute uniformly. Systems that
implement consolidation almost always also implement forgetting (14 of 16, or
87.5\%), but the reverse is not true---many systems with forgetting (34 total)
lack consolidation. Similarly, graph structure strongly predicts hybrid
retrieval fusion: 12 of 14 systems with fusion also have graph traversal.
These correlations suggest that certain capabilities are architecturally
co-dependent: consolidation requires a semantic store to consolidate
\emph{into}, which in turn motivates forgetting to manage the growing semantic
store; graph traversal provides a third retrieval signal that motivates
principled fusion to combine it with sparse and dense signals.

Third, the long tail creates an \textbf{evaluation gap}. Benchmarks like
LoCoMo and LongMemEval evaluate memory recall and factual accuracy, but they
do not test forgetting quality, preference tracking, contradiction resolution,
or associative retrieval through graph structure. As a result, simple systems
that score well on retrieval benchmarks may be indistinguishable from
architecturally richer systems in published evaluations, obscuring the
practical benefits of lifecycle operations that only manifest over extended
use. The field needs benchmarks that evaluate memory \emph{management}, not
just memory \emph{retrieval}---a point we return to in
\S\ref{sec:open-problems}.
