\section{The MEMORY.md Problem}
\label{sec:memorymd}

File-based memory---MEMORY.md files, daily logs, and workspace
documents injected into the system prompt---works for single sessions
but degrades as usage grows.  The AI agent community has converged on
this diagnosis through independent experience~\cite{zilliz2026b,
chawla2026, fisher2026, azizi2026}.  This section documents four
failure modes (cost, structure, retrieval, extraction), catalogs the
practitioner workarounds that emerged in response, maps each workaround
to its equivalent in formal memory architecture, and distills five
convergent design principles.  The evidence is drawn from GitHub issues,
community discussions, blog posts, and open-source tools---practitioner
data not covered by existing surveys.

\subsection{The Cost Problem}
\label{sec:memorymd-cost}

OpenClaw, one of the most widely deployed AI coding agents, injects the
full contents of its workspace files (MEMORY.md, AGENTS.md, SOUL.md,
and related documents) into the system prompt on \emph{every} API call,
regardless of relevance.  A community analysis of token
consumption~\cite{openclaw9157} measured the impact: approximately
35,600 tokens per message in workspace injection alone.  In a typical
multi-message conversation, the average message comprises 38,016 tokens,
of which 35,600 are memory content---93.5\% of the token budget consumed
by context that is overwhelmingly irrelevant to the current query.

The financial consequences are substantial.  At current pricing of
\$15 per million output tokens, heavy users report costs of
\$3,600 per month attributable to memory injection
alone~\cite{zilliz2026b, azizi2026}.  A community discussion documented
real-world costs dropping from \$4.20/day to \$1.80/day after applying
workarounds---a 57\% reduction that underscores how much of the baseline
cost is pure waste.

Three community tools emerged independently to replace full-context
injection with ranked retrieval:

\begin{table}[!t]
\centering
\caption{Community tools replacing full-context memory injection with
ranked retrieval.  All three independently adopted Reciprocal Rank
Fusion (RRF)~\cite{cormack2009rrf} for score combination.}
\label{tab:community-retrieval-tools}
\small
\begin{tabular}{@{}L{2.2cm}L{2.4cm}C{1.2cm}@{}}
\toprule
\textbf{Tool} & \textbf{Retrieval Pipeline} & \textbf{Token Savings} \\
\midrule
QMD~\cite{lutke2026qmd, casanova2026}
  & BM25 + vector + RRF + local cross-encoder reranking
  & 96\%~\cite{levine2026} \\
\addlinespace
memsearch~\cite{zilliz2026a}
  & BM25 + vector + RRF with top-$k$ progressive disclosure
  & Not quantified \\
\addlinespace
Clawdbot-Next TGAA~\cite{azizi2026}
  & Tiered Global Anchor Architecture + dynamic tool injection
  & 70--90\% \\
\bottomrule
\end{tabular}
\end{table}

A striking pattern emerges from Table~\ref{tab:community-retrieval-tools}:
all three tools independently adopted Reciprocal Rank
Fusion~\cite{cormack2009rrf} over the default weighted linear
combination used by the host platform.  This convergence is not
coincidental.  RRF combines ranked lists from heterogeneous retrievers
without requiring score normalization---a critical advantage when
merging BM25 scores (unbounded) with vector cosine similarities
(bounded to $[-1, 1]$)~\cite{qdrant2025, weaviate2024}.
Bruch~\etal~\cite{bruch2023} demonstrated that convex combination
outperforms RRF when labeled tuning data is available, but RRF is
superior in zero-shot settings.  Agent memory is characteristically a
zero-shot domain: labeled relevance judgments do not exist, and query
distributions shift with each session.  The practitioner community
arrived at this conclusion empirically, and it aligns with the same
fusion strategy used by production systems such as
Zep/Graphiti~\cite{dempsey2025graphiti} and multiple dedicated memory
engines surveyed in~\S\ref{sec:systems}.

\subsection{The Structure Problem}
\label{sec:memorymd-structure}

MEMORY.md conflates decisions, preferences, factual knowledge, and
session context into a single unstructured file.  As the file grows,
retrieval degrades (everything is a candidate for every query), and the
agent cannot distinguish a finalized architectural decision from a
speculative note.  Users have independently invented structured memory
patterns to work around these limitations.
Table~\ref{tab:structure-workarounds} catalogs six representative
patterns and maps each to its formal memory system equivalent.

\begin{table*}[!t]
\centering
\caption{Practitioner-invented structured memory patterns and their
equivalents in formal memory architectures.  Each workaround addresses a
specific failure of flat-file memory.  The ``System Equivalent'' column
uses generic architectural terms, not any specific implementation.}
\label{tab:structure-workarounds}
\small
\begin{tabular}{@{}L{3.4cm}L{2.5cm}L{3.7cm}L{3.8cm}@{}}
\toprule
\textbf{Pattern} & \textbf{Author} & \textbf{Description} & \textbf{System Equivalent} \\
\midrule
\texttt{decision.md}
  & Genet~\cite{genet2026}
  & Separate file for finalized decisions; user explicitly tells the
    agent ``that's a decision'' to prevent re-litigation across sessions.
  & Semantic node with a \emph{fact} type.  Consolidation promotes the
    decision from episodic history to the persistent semantic store. \\
\addlinespace
\texttt{working-context.md}
  & Fisher~\cite{fisher2026}
  & Real-time context snapshot updated during complex tasks; designed to
    survive context-window compaction.
  & Episodic store with session identifiers.  Full session context
    persists in the database, immune to context-window truncation. \\
\addlinespace
Temporal metadata (\texttt{[since:]}, \texttt{[learned:]})
  & Fisher~\cite{fisher2026}
  & Bi-temporal markers tracking when a fact became true versus when the
    system learned it (inspired by Graphiti~\cite{dempsey2025graphiti}).
  & \texttt{created\_at} and \texttt{last\_corroborated} timestamps on
    semantic nodes, enabling provenance queries and temporal filtering. \\
\addlinespace
4-layer architecture (SOUL.md + tacit.md + daily notes + PARA vault)
  & Oblivion Labz
  & Identity layer, tacit knowledge layer, temporal notes, and a
    reference vault---all prepended to the system prompt.
  & Multi-store architecture: separate episodic, semantic, and
    procedural stores with distinct retention policies. \\
\addlinespace
12-layer architecture with activation/decay
  & coolmanns~\cite{coolmanns2026}
  & SQLite + FTS5 knowledge graph with Hot/Warm/Cool activation tiers
    and cron-based nightly decay.
  & Forgetting model with tiered retrieval strength.  Activation decays
    over time; frequently accessed memories retain high accessibility. \\
\addlinespace
TELOS identity system (10 structured files)
  & Miessler, adapted by Swift
  & Values, goals, and background as structured Markdown in an
    Obsidian vault, organized by semantic category.
  & Typed semantic store with category-specific schemas, separating
    identity facts from task knowledge. \\
\bottomrule
\end{tabular}
\end{table*}

Several observations follow from Table~\ref{tab:structure-workarounds}.
First, the community has reinvented the episodic--semantic distinction
independently.  Genet's \texttt{decision.md} separates curated facts
from raw conversation history, precisely mirroring the consolidation
pathway from episodic to semantic memory described in Complementary
Learning Systems (CLS) theory~\cite{mcclelland1995}.  Fisher's temporal
metadata recreates the bi-temporal provenance tracking that
Zep/Graphiti~\cite{dempsey2025graphiti} implements at the graph layer.

Second, multi-layer architectures (4-layer and 12-layer) recapitulate
the multi-store models of cognitive psychology.  The 12-layer system is
particularly notable: its Hot/Warm/Cool activation tiers implement a
form of retrieval-strength decay, the same principle formalized in the
Bjork dual-strength model~\cite{bjork1992}, where retrieval strength
decays over time while storage strength persists.

Third, these workarounds remain fundamentally constrained by the
flat-file medium.  They impose structure through naming conventions and
manual discipline, not through schema enforcement or automated
lifecycle management.  A user who forgets to tag a fact with
\texttt{[since:]} loses its temporal provenance permanently.  A formal
memory system encodes these distinctions in the schema itself, making
them mandatory rather than optional.

\subsection{The Retrieval Problem}
\label{sec:memorymd-retrieval}

Chawla~\cite{chawla2026} demonstrated the canonical retrieval failure
of flat-file memory systems.  Consider the following scenario: on Monday,
the user mentions ``Alice manages the authentication team.''  On
Wednesday, the user asks ``who handles auth permissions?''  A vector
similarity search over the memory file retrieves entries containing
``authentication'' but ranks them alongside any other entry with a
similar embedding---including, in Chawla's example, notes about
password salting that happen to share embedding-space proximity with the
query.  More critically, even when the correct episode is retrieved, the
system cannot connect the entities: it knows Alice exists and auth
exists, but the relationship between them is implicit in the text rather
than explicit in the data model.

Three architectural mechanisms address this failure, none of which are
available to flat-file memory:

\begin{enumerate}[leftmargin=*, nosep]
    \item \textbf{Typed memory nodes.}  When the consolidation process
    extracts ``Alice manages the authentication team'' as a semantic
    node of type \emph{relationship}, the retrieval system can restrict
    search to relationship nodes when answering a ``who handles X?''
    query.  This narrows the candidate set, eliminating irrelevant
    matches that vector similarity alone cannot distinguish.

    \item \textbf{Graph links.}  If the system maintains a graph
    overlay where entities (Alice, authentication, permissions) are
    connected by typed edges, then retrieving ``authentication'' also
    surfaces ``Alice'' through edge traversal---regardless of whether
    their text embeddings are close in vector space.  This is the
    approach adopted by knowledge-graph memory systems such as
    Graphiti~\cite{dempsey2025graphiti} and HippoRAG~\cite{gutierrez2024hipporag}.

    \item \textbf{Spreading activation.}  Collins and
    Loftus~\cite{collins1975} showed that activating a concept in a
    semantic network propagates activation to associated concepts through
    weighted edges.  Applied to agent memory, activating ``auth'' in the
    graph raises the activation level of ``Alice'' through the connecting
    edge, even when the user has not mentioned Alice explicitly.  This
    mechanism explains why associative retrieval outperforms pure
    embedding similarity for relational queries: the graph captures
    relationships that the embedding space does not.
\end{enumerate}

The community has recognized this limitation.  A feature request for
a next-generation memory system proposed associative traversal,
salience weighting, and access-based forgetting---mechanisms that
correspond to spreading activation, retrieval-strength scoring, and
dual-strength decay, respectively.  The convergence between practitioner
intuition and established cognitive science models is notable: users
who have never encountered Collins and Loftus independently arrive at
the need for activation propagation through associative links.

\subsection{The Extraction Problem}
\label{sec:memorymd-extraction}

The most fundamental limitation of file-based memory is that the user
must manually curate it.  There is no automated pathway from
conversation to structured knowledge.  Fisher~\cite{fisher2026}
observed that ``the extraction step is more important than the storage
format''---getting facts out of conversations automatically is the
transformative capability.

Fisher's own approach addresses extraction partially: temporal metadata
and manual organization impose structure on the memory file, but the
human remains the extraction engine.  This does not scale.  A user who
conducts 20 sessions per week cannot reasonably curate the resulting
knowledge base by hand.

The automated alternative is \emph{consolidation}: a process inspired
by CLS theory~\cite{mcclelland1995} in which a provider examines raw
episodic records (conversation transcripts) and extracts structured
semantic nodes (facts, relationships, preferences).  This mirrors the
hippocampus-to-neocortex replay that McClelland~\etal\ described:
episodes are replayed during a consolidation phase, and stable patterns
are promoted to long-term semantic storage while transient details are
discarded.

In the memory architecture context, consolidation providers typically
implement three operations: \emph{knowledge extraction} (identifying
facts and relationships in episodes), \emph{impression extraction}
(identifying behavioral patterns and preferences), and
\emph{contradiction detection} (identifying conflicts between new
extractions and existing semantic nodes).  Making extraction a
first-class architectural concern---rather than an afterthought bolted
onto a file-based system---ensures that the memory store grows
automatically as the agent interacts.

Early prototype results for CLS-inspired extraction are instructive.
A community prototype achieved F1$\approx$44\% on a 45-query
benchmark, with recall at 65\% but precision at only 35\%.  The
asymmetry is significant: 65\% of retrieved content was noise.  For
memory systems, \textbf{precision matters more than recall}.  A
forgotten fact can be relearned from a future conversation, but a false
memory---a confidently asserted ``fact'' that was never stated---can
propagate errors through subsequent reasoning chains.  This asymmetry
suggests that extraction systems should be calibrated conservatively,
preferring to miss genuine facts over fabricating spurious ones.

\subsection{Convergent Design Principles}
\label{sec:memorymd-convergence}

The practitioner community---working independently, without
coordination, and largely without awareness of the cognitive science
literature---has converged on five architectural principles for agent
memory:

\begin{enumerate}[leftmargin=*, nosep]
    \item \textbf{Ranked retrieval over full-context injection.}  The
    tools in Table~\ref{tab:community-retrieval-tools} demonstrate that
    selectively retrieving relevant memories (via BM25 + vector + RRF)
    reduces token consumption by 70--96\% with no loss in agent
    performance.  Full-context injection is the memory equivalent of
    reading the entire encyclopedia to answer a single question.

    \item \textbf{Typed memory stores over flat files.}  The patterns
    in Table~\ref{tab:structure-workarounds} show users independently
    creating separate files for decisions, context, identity, and
    knowledge---reinventing the episodic/semantic/procedural distinction
    from cognitive psychology.

    \item \textbf{Graph traversal over pure text similarity.}  The
    retrieval failures documented by Chawla~\cite{chawla2026} drove
    proposals for knowledge-graph overlays, entity linking, and
    associative traversal---mechanisms that require structured
    relationships rather than flat text.

    \item \textbf{Temporal decay over manual curation.}  The 12-layer
    architecture~\cite{coolmanns2026} with Hot/Warm/Cool activation
    tiers and Fisher's~\cite{fisher2026} temporal metadata both
    recognize that memories have a time dimension: recent, frequently
    accessed knowledge should outrank stale entries.  Manual curation
    does not scale; automated decay does.

    \item \textbf{Automatic extraction over manual logging.}  Fisher's
    observation that extraction is ``more important than the storage
    format''~\cite{fisher2026} reflects a community-wide recognition
    that the bottleneck is not \emph{where} to store memories but
    \emph{how} to create them.  Manual logging requires discipline that
    degrades over time; automated consolidation operates consistently.
\end{enumerate}

These five principles are not arbitrary.  Each corresponds to an
established result in cognitive neuroscience or information retrieval:
ranked retrieval maps to the selective attention and cue-dependent
recall studied in memory psychology; typed stores map to the
episodic--semantic distinction central to CLS
theory~\cite{mcclelland1995}; graph traversal maps to spreading
activation in semantic networks~\cite{collins1975}; temporal decay
maps to the retrieval-strength component of the Bjork dual-strength
model~\cite{bjork1992}; and automatic extraction maps to the
hippocampal replay mechanism in complementary learning
systems~\cite{mcclelland1995}.

The Yogac\=ara Buddhist tradition, discussed in
\S\ref{sec:yogacara}, arrives at structurally parallel insights through
a different analytical path.  The \emph{b\=ija} (seeds) stored in the
\emph{\=alaya-vij\~n\=ana} (storehouse consciousness) are strengthened
through \emph{v\=asan\=a} (perfuming)---a process that maps to
Hebbian potentiation of co-activated memory traces.  Seeds that are not
perfumed weaken over time, corresponding to retrieval-strength decay.
The storehouse itself is not a static repository but a dynamic process
that reshapes through interaction (\emph{\=a\'sraya-par\=av\d{r}tti},
basis-transformation).

The convergence across three independent traditions---practitioner
intuition, cognitive neuroscience, and contemplative
psychology---suggests that these five principles are not design
preferences but \emph{structural necessities} of any sufficiently
complex memory system.  A system that stores everything retrieves
nothing useful (violating principle~1).  A system that conflates memory
types cannot selectively query (violating principle~2).  A system
without relational structure cannot answer relational queries (violating
principle~3).  A system without temporal dynamics drowns in stale
information (violating principle~4).  A system without automated
extraction cannot scale beyond manual effort (violating principle~5).
The practitioner community discovered these constraints the hard way,
through the accumulated friction of daily use.  That their solutions
align with both neuroscience and a 1,500-year-old contemplative
tradition suggests the constraints are fundamental.
