\section{A Yogac\=ara Framework for Memory Taxonomy}
\label{sec:yogacara}

The Yogac\=ara (\cjk{瑜伽行派}) school of Buddhist philosophy,
systematized in the \emph{Yogac\=arabh\=umi-\\'s\=astra}
(\cjk{瑜伽師地論}, attributed to Maitreya/Asa\.{n}ga, translated by
Xuanzang 646--648~CE)~\cite{yogacarabhumi} and the \emph{Cheng Weishi
Lun} (\cjk{成唯識論}, Xuanzang, 659~CE)~\cite{chengweishilun},
developed a detailed phenomenology of memory more than a millennium
before computational cognitive science.  This section does not argue
that AI architects \emph{should} consult Buddhist texts.  Rather, we
demonstrate that the Yogac\=ara eight-consciousness model independently
arrived at structural distinctions that the AI memory community has
rediscovered---and in several cases, the Yogac\=ara framework
\emph{predicts} design choices that most production systems have not yet
adopted.

\subsection{Structural Parallels}
\label{sec:yogacara-structural}

The Yogac\=ara model posits eight consciousnesses
(\cjk{八識}), of which three are directly relevant to memory
architecture.  The \emph{\=alaya-vij\~n\=ana} (\cjk{阿賴耶識},
storehouse consciousness) is a persistent substrate that holds all
experiential potentials as \emph{b\=ija} (\cjk{種子}, seeds).  It has
no agency of its own; it stores and yields potentials when conditions
converge.  The \emph{Sa\d{m}dhinirmocana S\=utra}
(\cjk{解深密經})~\cite{samdhinirmocana} describes it as ``exceedingly
deep and subtle; all its seeds like a torrential flood''---a continuous
stream that underlies all mental activity.  Above it, \emph{manas}
(\cjk{末那識}, the seventh consciousness) performs deliberative
reflection, continuously examining and reinterpreting what the
storehouse contains.

\Cref{tab:yogacara-mapping} maps six Yogac\=ara concepts to the
four-axis taxonomy introduced in \Cref{sec:taxonomy}.  Three
observations make the mapping non-trivial.

\begin{table}[t]
\centering
\caption{Mapping Yogac\=ara Concepts to Memory Architecture}
\label{tab:yogacara-mapping}
\begin{tabular}{@{}L{2.2cm}L{2.2cm}L{3cm}@{}}
\toprule
\textbf{Yogac\=ara} & \textbf{Memory Axis} & \textbf{AI Implementation} \\
\midrule
\emph{\=Alaya-vij\~n\=ana} & Structure & Persistent memory store \\
\emph{B\=ija} (seeds) & Structure & Memory nodes with strengths \\
\emph{V\=asan\=a} & Preference & Impression accumulation \\
\emph{Vip\=aka} & Lifecycle & Consolidation (heterogeneous ripening) \\
\emph{\=A\'sraya-par\=av\d{r}tti} & Lifecycle & Transformation (dedup, prune, resolve) \\
\emph{Vijnapti-m\=atra} & Retrieval & Encoding specificity \\
\bottomrule
\end{tabular}
\end{table}

First, b\=ija are not static records.  The
\emph{Yogac\=arabh\=umi-\\'s\=astra} describes seeds as possessing
variable strengths: seeds that are repeatedly activated through
experience grow stronger, while those left unactivated weaken.  This
directly parallels the dual-strength model of Bjork and
Bjork~\cite{bjork1992}, where every memory node carries independent
storage and retrieval strengths that change with use.

Second, the relationship between \=alaya-vij\~n\=ana and manas mirrors
the CLS distinction between a fast-encoding episodic store and a
slow-learning semantic store~\cite{mcclelland1995}.  The storehouse
receives impressions rapidly and indiscriminately; manas reflects on
those impressions deliberatively, producing structured understanding
over time.  In AI memory architectures, this corresponds to the
episodic--semantic separation found in systems such as Generative
Agents~\cite{park2023generative}, SYNAPSE~\cite{jiang2026synapse}, and
\alaya{}.

Third, the Yogac\=ara model treats memory as inherently
\emph{processual}.  Seeds do not simply sit in the storehouse; they
ripen, perfume, condition one another, and periodically undergo
structural transformation.  This emphasis on active lifecycle operations
aligns with the central argument of our taxonomy: that formation and
retrieval alone are insufficient---consolidation, transformation, and
forgetting are equally essential axes of memory architecture design.

\subsection{Predictive Power}
\label{sec:yogacara-predictive}

We identify four cases in which the Yogac\=ara framework describes a
structural principle that the AI agent memory community has either
independently converged upon or has not yet widely adopted.  In each
case, the Yogac\=ara claim is not metaphorical; it is a specific,
falsifiable assertion about how a memory system should behave.

\paragraph{Heterogeneous ripening (\emph{vip\=aka}, \cjk{異熟})}
The \emph{Cheng Weishi Lun}~\cite{chengweishilun} insists that seeds
ripen into manifest experience that \emph{differs in nature from the
cause}: moral seeds ripen into pleasure or pain, not into further
morality.  The term \cjk{異熟} literally means ``differently
ripened.''  Translated into architectural terms, this predicts that
consolidation should produce memory nodes qualitatively different from
the episodes that generated them---decontextualized, restructured, and
reusable in novel contexts.  This is precisely what Complementary
Learning Systems theory formalized in 1995~\cite{mcclelland1995}:
hippocampal replays produce neocortical representations that differ in
kind from the original episodes.  Among the systems we survey,
those that implement true consolidation (Generative Agents,
SYNAPSE, HippoRAG~\cite{gutierrez2024hipporag}, \alaya{}) produce
semantic nodes structurally distinct from their episodic sources.  The
many systems that store only raw episodes or only extracted facts
implicitly assume homogeneous ripening---an assumption the Yogac\=ara
model would reject.

\paragraph{Gradual perfuming (\emph{v\=asan\=a}, \cjk{薰習})}
Each experience leaves a subtle trace on the storehouse ``like incense
gradually permeating cloth''~\cite{yogacarabhumi}.  No single trace is
transformative; preferences emerge from the accumulation of many similar
impressions.  This predicts that a memory system should derive user
preferences from \emph{accumulated} observations, not from one-shot
extraction.  Most production systems (Mem0~\cite{choudhary2025mem0},
A-MEM~\cite{xu2025amem}) extract preferences at write time through a
single LLM call per interaction.  These systems conflate signal with
noise: a user mentioning Python once is treated identically to a user
who has discussed Python across fifty sessions.  The Yogac\=ara model
predicts that systems employing gradual accumulation with a
crystallization threshold---as \alaya{} does with its five-impression
minimum before preference formation---will produce more stable and
accurate preference representations.

\paragraph{Transformation of the basis (\emph{\=a\'sraya-par\=av\d{r}tti},
\cjk{轉依})}
Yogac\=ara describes a periodic structural transformation of the
storehouse itself: not merely appending new content, but resolving
contradictions, pruning obsolete potentials, and restructuring what
remains toward greater clarity.  The
\emph{Yogac\=arabh\=umi-\\'s\=astra}~\cite{yogacarabhumi} describes
the transformation of the eight consciousnesses into the four wisdoms
(\cjk{轉識成智}), of which the first is ``Great Mirror Wisdom''
(\cjk{大圓鏡智})---reflecting reality without redundant distortion.
This predicts that memory systems require periodic reorganization
operations beyond append-only growth: deduplication, contradiction
resolution, and decay of unreinforced entries.  Most systems in our
survey are append-only; the few that implement transformation operations
(\alaya{}'s \texttt{transform()} with deduplication, link pruning, and
preference decay; Letta's~\cite{packer2023memgpt} sleep-time
consolidation) align with the Yogac\=ara prediction.  The absence of
transformation in most architectures is, from this perspective, a
design gap rather than a simplification.

\paragraph{Perspective-relativity (\emph{vijnapti-m\=atra}, \cjk{唯識})}
The Yogac\=ara doctrine of ``consciousness-only'' asserts that what is
stored is not an objective record of reality but a representation shaped
by the conditions of its formation.  Memory is always
\emph{perspective-relative}: the same event, encoded under different
conditions, yields different seeds.  This directly parallels Tulving and
Thomson's encoding specificity principle~\cite{tulving1973}: a memory is
most accessible when the retrieval context matches the encoding context.
In AI memory architectures, this predicts that storing rich contextual
metadata at encoding time (topics, sentiment, entities, conversational
role) and using that metadata at retrieval time will outperform
context-free similarity search.  Systems that store episode context and
use it for reranking---such as Generative Agents (recency $\times$
importance $\times$ relevance) and \alaya{} (topic overlap, entity
overlap, sentiment proximity)---instantiate this principle.

\subsection{Convergence with Neuroscience}
\label{sec:yogacara-neuroscience}

The parallels between Yogac\=ara and cognitive neuroscience are
especially striking because the two traditions developed independently,
separated by fifteen centuries and entirely different methodological
commitments.  Yogac\=ara relied on introspective phenomenology and
philosophical analysis; neuroscience relies on experimental observation
and computational modeling.  Their convergence on the same structural
patterns provides independent triangulation for the design principles
underlying our taxonomy.

\paragraph{Dual-system consolidation}
CLS theory~\cite{mcclelland1995} proposes that the brain separates a
fast-learning hippocampal system (episodic encoding) from a
slow-learning neocortical system (semantic integration), with
interleaved replay during sleep transferring knowledge from the former
to the latter.  The Yogac\=ara distinction between \=alaya-vij\~n\=ana
(which receives all impressions rapidly) and manas (which reflects on
them deliberatively) describes the same architectural split at the
phenomenological level.  Both frameworks predict that a memory
architecture combining rapid, high-fidelity encoding with gradual,
lossy extraction of stable patterns will outperform either component
alone.

\paragraph{Hebbian strengthening and weakening}
Hebb's principle~\cite{hebb1949}---``neurons that fire together wire
together''---and its physiological substrate in long-term potentiation
(LTP) and long-term depression (LTD)~\cite{bliss1973} describe a
mechanism by which synaptic connections strengthen through co-activation
and weaken through disuse.  The \emph{Yogac\=arabh\=umi-\\'s\=astra}
describes seeds as strengthening through repeated activation (\cjk{增長})
and weakening when not activated (\cjk{損減})---the same dynamics
expressed in phenomenological rather than physiological vocabulary.
In both frameworks, the strength of a stored association is a function
of its history of activation, not a fixed property assigned at encoding
time.

\paragraph{Spreading activation and seed mutual conditioning}
Collins and Loftus~\cite{collins1975} formalized spreading activation in
semantic networks: activating one concept propagates activation to
associated concepts through weighted links, with decay proportional to
graph distance.  The
\emph{Yogac\=arabh\=umi-\\'s\=astra}~\cite{yogacarabhumi} describes
seeds as mutually conditioning (\cjk{互相薰習})---the ripening of one
seed creates conditions for the ripening of associated seeds.  The
structural isomorphism is precise: in both models, (1)~stored elements
are interconnected, (2)~activation of one element facilitates activation
of neighbors, (3)~the degree of facilitation is proportional to
connection strength, and (4)~facilitation attenuates with distance.
This convergence suggests that associative, graph-structured retrieval
is not merely a useful engineering heuristic but a fundamental
organizational principle for memory systems.

\paragraph{Competitive dynamics and retrieval-induced forgetting}
Anderson, Bjork, and Bjork~\cite{anderson1994} demonstrated that
retrieving a memory actively suppresses competing memories sharing the
same retrieval cues---a phenomenon termed retrieval-induced forgetting
(RIF).  Yogac\=ara describes an analogous competitive dynamic among
seeds: when conditions favor the ripening of certain seeds, competing
seeds (\cjk{相違種子}, ``mutually opposed seeds'') are
suppressed~\cite{chengweishilun}.  Both frameworks predict that
retrieval is not a passive lookup but an active process that reshapes
the accessibility landscape of the memory store.  In AI memory
architectures, this implies that retrieval operations should update
memory strengths---boosting accessed nodes and allowing competing nodes
to decay---rather than leaving the store unchanged after a query.

\medskip

\noindent These convergences do not establish causal or historical
influence.  Rather, they suggest that Yogac\=ara phenomenology,
cognitive neuroscience, and AI memory engineering are independently
constraining the same design space.  When three traditions separated by
methodology, culture, and fifteen centuries arrive at the same
structural principles---dual-system consolidation, use-dependent
strengthening, associative spreading, and competitive retrieval
dynamics---those principles merit attention as candidate universals of
memory architecture.
