\section{Open Problems and Future Directions}
\label{sec:open-problems}

The preceding analysis reveals that while memory architectures for
LLM-based agents have proliferated rapidly, several fundamental
problems remain underexplored. We identify seven open research
directions that, if addressed, would substantially advance the field
from static memory stores toward memory as a living process.

\subsection{Principled Forgetting}

The majority of surveyed systems implement no forgetting mechanism
whatsoever, treating memory as an append-only store. Among those that
do, most rely on single-curve decay models: MemoryBank and
CortexGraph apply Ebbinghaus forgetting
curves~\cite{ebbinghaus1885}, Mem0 uses exponential decay, and
Generative Agents~\cite{park2023generative} weight memories by
recency. SYNAPSE~\cite{jiang2026synapse} introduces temporal decay
with lateral inhibition, providing competitive suppression beyond
simple time-based fading. The most principled approach is the Bjork
dual-strength model~\cite{bjork1992}, which distinguishes storage
strength (how deeply encoded) from retrieval strength (how currently
accessible), combined with retrieval-induced
forgetting~\cite{anderson1994} in which retrieving one memory actively
suppresses competitors sharing the same cues. Despite this diversity
of approaches, no empirical
comparison of forgetting models exists on standard agent benchmarks
such as LoCoMo or LongMemEval. Three critical questions remain
unanswered: whether principled forgetting improves downstream task
performance beyond what unlimited storage achieves, whether forgetting
functions better as a retrieval quality mechanism than as a storage
management strategy, and which forgetting model best balances memory
fidelity against retrieval precision across interaction horizons
spanning weeks to years.

\subsection{Preference Emergence vs.\ Extraction}

Two fundamentally different paradigms exist for modeling user
preferences. The dominant approach---exemplified by
Mem0~\cite{choudhary2025mem0} and Supermemory---uses LLM-based
extraction, invoking the language model on each interaction to
identify and store explicit preference statements. This approach is
fast and produces interpretable outputs but is brittle (preferences
that are implied rather than stated are missed), expensive (each
extraction consumes inference tokens), and prone to recency bias.
The alternative is emergent crystallization: individual impressions
accumulate over many interactions, and preferences crystallize
organically when sufficient evidence converges, without any
per-interaction LLM call. Hindsight~\cite{latimer2025hindsight}
occupies a middle ground with its opinion memory, which maintains
confidence-scored belief evolution that tracks how user preferences
shift over time. No existing benchmark evaluates preference modeling
accuracy---neither extraction fidelity nor emergence quality. The
central research question is when preferences require explicit
extraction (e.g., dietary restrictions, access credentials) versus
when they are better served by organic emergence (e.g., communication
style, aesthetic sensibilities), and whether hybrid approaches can
combine the immediacy of extraction with the robustness of
accumulation.

\subsection{Cross-Memory Consolidation}

Most memory systems store all information in a single flat structure,
whether a vector database or a knowledge graph. Episodic-to-semantic
consolidation---the process by which raw conversational episodes are
distilled into stable knowledge representations---remains rare
despite strong neuroscientific motivation. Systems that implement
consolidation include Generative
Agents~\cite{park2023generative} (periodic reflection that
synthesizes higher-order observations),
SYNAPSE~\cite{jiang2026synapse} (activation-based promotion from
episodic to semantic stores),
LightMem~\cite{fang2025lightmem} (``sleep-time'' background
processing), and EverMemOS (an encoding-consolidation-retrieval
pipeline modeled on memory systems neuroscience). Yet even among
these systems, consolidation policies are hand-crafted: fixed batch
sizes, static frequency schedules, and threshold-based triggers.
Open questions include determining the optimal consolidation frequency
(too aggressive loses episodic detail; too conservative accumulates
redundant raw data), defining quality metrics for consolidated
representations (is a semantic node ``better'' than the episodes it
summarizes, and by what measure?), and whether consolidation policies
should be learned via reinforcement learning rather than
hand-engineered---an approach that the RL-trained memory management
trend (\S\ref{sec:temporal-trends}) suggests is viable but that no
system has yet applied specifically to consolidation.

\subsection{Adaptive Retrieval}

Retrieval pipelines in most surveyed systems are static: a fixed
sequence of embedding lookup, optional reranking, and top-$k$
truncation that does not change regardless of how many queries the
system processes. Three alternatives have emerged. First, Hebbian
graph approaches---implemented in
SYNAPSE~\cite{jiang2026synapse} and similar systems---reshape associative pathways
through use: co-retrieved memories strengthen their mutual links
(LTP), while unused associations decay (LTD), producing emergent
small-world topology without explicit supervision. Second, RL-learned
retrieval policies---pursued by
Mem-alpha~\cite{wang2025memalpha},
MemRL~\cite{zhang2026memrl}, and Memory-R1---train the memory
management policy end-to-end from task reward signals. Third,
graph-adaptive algorithms such as HippoRAG's Personalized
PageRank~\cite{gutierrez2024hipporag} bias traversal toward
query-relevant subgraphs but do not adapt based on historical usage
patterns. Collins and Loftus's spreading activation
theory~\cite{collins1975} predicts that retrieval should reshape the
network it traverses, yet only Hebbian approaches implement this
bidirectional coupling. Whether RL-learned policies can match the
organic adaptivity of Hebbian graphs---or whether hybrid approaches
combining learned policies with usage-driven graph evolution offer
superior performance---remains an open empirical question.

\subsection{Multi-Agent Memory}

The vast majority of memory systems assume a single-agent setting, yet
production deployments increasingly involve multiple agents
collaborating on shared tasks. G-Memory~\cite{zhang2025gmemory}
introduces a three-tier hierarchy---agent-local, team-shared, and
global memory---with explicit ownership tracking and conflict
resolution protocols, representing the most structured approach to
date. However, several fundamental problems remain unresolved. Memory
ownership semantics are unclear: when Agent~A consolidates an episode
from a conversation that Agent~B also participated in, which agent
owns the resulting semantic node? Conflict resolution across agents
with different memory backends (e.g., one using a vector store,
another a knowledge graph) requires alignment mechanisms that no
current system provides. Privacy boundaries introduce additional
complexity: in multi-tenant deployments, agents serving different
users must maintain strict memory isolation while potentially sharing
task-relevant procedural knowledge. Federated memory architectures,
in which agents share consolidated knowledge without exposing raw
episodic data, represent a promising but entirely unexplored research
direction that would bridge the gap between single-agent memory
systems and the collaborative agent ecosystems that production
environments demand.

\subsection{Evaluation Methodology}

The first generation of memory benchmarks---LoCoMo, LongMemEval, and
DMR---evaluate memory systems primarily on retrieval accuracy: given a
query about past interactions, does the system return the correct
information?  While valuable, these benchmarks treat memory as a static
store and fail to evaluate the lifecycle processes that distinguish
sophisticated memory architectures from simple retrieval systems.

Recent benchmarks have begun to close specific gaps.
TReMu~\cite{ge2025tremu} targets temporal reasoning,
LoCoMo-Plus~\cite{li2026locomoplus} evaluates cognitive memory under
cue--trigger semantic disconnect, and
MemoryAgentBench~\cite{hu2025mab} explicitly tests selective forgetting
and conflict resolution alongside retrieval and learning
(\S\ref{sec:benchmark-landscape}).  Yet even collectively, these
benchmarks leave critical dimensions untested: no existing benchmark
measures consolidation quality (are semantic nodes more useful for
downstream tasks than the raw episodes they replace?), preference
modeling accuracy (does the system correctly infer and apply user
preferences over extended interaction horizons?), or the interaction
between forgetting and retrieval precision over time.  Furthermore,
all current benchmarks operate on fixed histories, precluding evaluation
of how memory evolves under continuous interaction.

We argue that the field requires process-oriented benchmarks that
evaluate memory as a dynamic system: measuring not only what is
retrieved but how retrieval changes over time, whether forgetting
improves signal-to-noise ratio, and whether consolidation produces
representations that generalize better than raw storage.  Such
benchmarks would need to simulate multi-session interactions spanning
weeks or months, with evaluation criteria that capture the trajectory
of memory quality rather than point-in-time accuracy alone.

\subsection{Ethical Considerations}

Memory architectures for LLM agents raise ethical questions that the
field has largely deferred. The right to be forgotten, codified in
GDPR Article~17, requires that personal data be erasable upon
request---yet memory systems with consolidation, graph links, and
derived preferences make complete erasure technically challenging, as
a single episode may have influenced semantic nodes, graph weights,
and crystallized preferences in ways that are difficult to trace.
Memory manipulation through adversarial injection of false memories
poses integrity risks: if an attacker can insert fabricated episodes,
downstream consolidation may propagate falsehoods into the system's
stable knowledge. Bias amplification through accumulation mechanisms
is particularly concerning: any accumulation mechanism in which
preferences crystallize from repeated impressions can entrench
biases just as readily as it captures genuine preferences. Finally, if memory is
understood as a process rather than a store---continuously
consolidating, forgetting, adapting, and crystallizing---the question
of when such a process constitutes something resembling a ``self''
moves from philosophical abstraction to engineering concern.
Addressing these questions requires interdisciplinary collaboration
among AI researchers, bioethicists, legal scholars, and philosophers
of mind.
