\section{Implementation}
\label{sec:implementation}

\alaya{} is implemented in Rust as an embeddable library crate, prioritizing
zero-dependency deployment and deterministic behavior.

\subsection{Storage Layer}

All data resides in a single SQLite database using WAL (Write-Ahead Logging)
mode for concurrent read access during write operations. The schema comprises
eight tables: \texttt{episodes} (with FTS5 virtual table for full-text
search), \texttt{semantic\_nodes}, \texttt{impressions},
\texttt{preferences}, \texttt{embeddings}, \texttt{links} (the Hebbian
graph), \texttt{node\_strengths} (Bjork dual-strength values),
\texttt{categories} (emergent ontology, with optional \texttt{parent\_id}
for hierarchical nesting), and \texttt{tombstones} (cascade deletion audit
trail, recording node type, ID, deletion timestamp, and reason for each
purged node). Schema versioning uses SQLite's
\texttt{PRAGMA user\_version} (currently version~4) with forward migration
on database open.

The choice of SQLite over a dedicated vector database or graph database is
deliberate. The companion survey~\cite{hui2026taxonomy} reports that 70.5\%
of memory systems require an LLM for memory management operations and 42.0\%
require 1--2 external services. \alaya{} requires neither: all memory
management operations (consolidation, transformation, forgetting,
perfuming, ontology) execute locally against SQLite, with the LLM used only
during consolidation for knowledge extraction---and even that is pluggable
via the \texttt{ConsolidationProvider} trait.

\subsection{Provider Interface}

\alaya{} does not embed any LLM or embedding model. Instead, it defines two
provider traits. The \texttt{ConsolidationProvider} trait handles knowledge
extraction, while the \texttt{EmbeddingProvider} trait (v0.2.0) provides
automatic embedding generation via \texttt{embed()} and
\texttt{embed\_batch()} methods, wired into \texttt{store\_episode()} and
\texttt{query()} so that applications need not manage vectors manually.
The \texttt{ConsolidationProvider} interface:

\begin{lstlisting}[caption={Provider trait (simplified)}]
pub trait ConsolidationProvider {
    fn extract_knowledge(
        &self,
        episodes: &[Episode],
    ) -> Result<Vec<NewSemanticNode>>;
}
\end{lstlisting}

This design keeps \alaya{} independent of any specific model, API, or
inference framework. Applications can use GPT-4, Claude, Gemini, local
models, or rule-based extractors. The emergent ontology system operates
entirely without the provider---category discovery uses embedding similarity
and graph topology, with LLM labeling as an optional enhancement.

\subsection{Embedding Handling}

Embeddings are stored as little-endian \texttt{f32} BLOBs in SQLite.
Cosine similarity is computed in Rust without external linear algebra
libraries. The system operates in three embedding modes:

\begin{itemize}[nosep]
    \item \textbf{Full embeddings:} Applications provide embeddings with
    episodes and semantic nodes. Vector retrieval and embedding-based
    clustering are fully operational.
    \item \textbf{Partial embeddings:} Some nodes have embeddings, others do
    not. The system uses embeddings where available and falls back to BM25
    and graph signals for nodes without embeddings.
    \item \textbf{No embeddings:} The system operates on BM25 full-text
    search and graph traversal alone. Category discovery falls back to
    graph-only clustering (shared Hebbian link overlap).
\end{itemize}

\subsection{MCP Server}

An optional Model Context Protocol (MCP)~\cite{anthropic2024mcp} server
binary (\texttt{alaya-mcp}) exposes \alaya{}'s API as MCP tools, enabling
any MCP-compatible AI agent to use \alaya{} as its memory backend. The MCP
server is gated behind a Cargo feature flag (\texttt{mcp}) and adds
\texttt{rmcp}, \texttt{tokio}, and \texttt{schemars} as dependencies only
when enabled. The default library crate has four dependencies:
\texttt{rusqlite} (with bundled SQLite), \texttt{serde},
\texttt{serde\_json}, and \texttt{thiserror}.

\subsection{Testing}

\alaya{} follows test-driven development with 232 tests (223 core + 9 MCP)
across four categories: deterministic unit tests covering individual store
operations, graph algorithms, lifecycle functions, category CRUD, hierarchy,
and splitting; property-based tests using \texttt{proptest} that verify
mathematical invariants (cosine similarity bounds, RRF score positivity and
ordering, Bjork strength clamping, Hebbian weight bounds, recency decay
monotonicity); integration tests covering full lifecycle pipelines (episode
ingestion through category emergence, including LTD decay, RIF suppression,
category splitting, and tombstone tracking); and documentation tests
ensuring code examples in the API documentation compile and produce correct
results. Continuous
integration runs the full test suite across three operating systems (Linux,
macOS, Windows) and two Rust toolchains (stable and MSRV 1.75).
