\section{Comparative Analysis}
\label{sec:comparison}

This section provides a quantitative cross-sectional analysis of the 116 memory
systems surveyed in \S\ref{sec:systems}, organized along three axes: feature
prevalence across the full landscape, temporal trends showing how the field has
evolved from 2023 to 2026, and the distribution of retrieval sophistication
mapped to the RAG taxonomy of Gao et al.~\cite{gao2023rag}. We exclude vector
databases (Pinecone, ChromaDB, Weaviate, Milvus, etc.) from all counts, as
these are infrastructure components rather than memory architectures. The
remaining 116 systems span five categories: Dedicated Memory Systems (38),
Framework Memory Modules (15), Coding Agent Memory (20), File-Based Memory (4),
and Research Architectures (19), plus 20 additional systems described in
sufficient detail to classify but not included in the primary comparison
tables.

%% ---------------------------------------------------------------
\subsection{Feature Prevalence}
\label{sec:feature-prevalence}

Table~\ref{tab:feature-prevalence} reports the count and percentage of systems
exhibiting each capability across the full survey. Features are organized into
six groups corresponding to the analytical framework introduced in
\S\ref{sec:background}: memory types drawn from the CoALA taxonomy
\cite{sumers2024coala}, retrieval signals mapped to the RAG survey
\cite{gao2023rag}, lifecycle operations from the memory operations taxonomy
\cite{zhang2024survey}, preference learning, graph structure, and
infrastructure requirements.

\begin{table}[!t]
\centering
\caption{Feature prevalence across 116 surveyed memory systems (excluding vector databases). Counts reflect systems with explicit support for each capability based on documentation, publications, and source code analysis.}
\label{tab:feature-prevalence}
\small
\begin{tabular}{@{}lrr@{}}
\toprule
\textbf{Feature} & \textbf{Count} & \textbf{\%} \\
\midrule
\multicolumn{3}{@{}l}{\textit{Memory Types (CoALA)}} \\
\quad Has episodic memory          & 64 & 57.1 \\
\quad Has semantic memory          & 58 & 51.8 \\
\quad Has procedural memory        &  8 &  7.1 \\
\quad Has implicit/preference mem. & 35 & 31.3 \\
\quad Has working memory           & 40 & 35.7 \\
\midrule
\multicolumn{3}{@{}l}{\textit{Retrieval Signals}} \\
\quad Uses BM25/sparse search      & 24 & 21.4 \\
\quad Uses vector/dense search     & 73 & 65.2 \\
\quad Uses graph traversal         & 32 & 28.6 \\
\quad Uses hybrid fusion (RRF or similar) & 19 & 17.0 \\
\quad Uses reranking               & 15 & 13.4 \\
\midrule
\multicolumn{3}{@{}l}{\textit{Lifecycle Operations}} \\
\quad Has any forgetting mechanism & 38 & 33.9 \\
\quad Has consolidation (ep.\ $\to$ sem.) & 19 & 17.0 \\
\quad Has contradiction resolution &  9 &  8.0 \\
\quad Has transformation/reorganization & 14 & 12.5 \\
\midrule
\multicolumn{3}{@{}l}{\textit{Preference Learning}} \\
\quad Has any preference learning  & 35 & 31.3 \\
\quad \quad via LLM extraction     & 21 & 18.8 \\
\quad \quad via emergent/accumulated & 5 &  4.5 \\
\midrule
\multicolumn{3}{@{}l}{\textit{Graph Structure}} \\
\quad Has any graph structure      & 32 & 28.6 \\
\quad \quad Static KG (LLM-built)  & 24 & 21.4 \\
\quad \quad Dynamic/Hebbian (usage-adaptive) & 4 & 3.6 \\
\midrule
\multicolumn{3}{@{}l}{\textit{Infrastructure Requirements}} \\
\quad Requires 0 external services & 35 & 31.3 \\
\quad Requires 1--2 external services & 43 & 38.4 \\
\quad Requires 3+ external services &  7 &  6.3 \\
\quad Requires LLM for memory ops  & 78 & 69.6 \\
\bottomrule
\end{tabular}
\end{table}

Three patterns emerge from Table~\ref{tab:feature-prevalence}. First, episodic
and semantic memory are roughly equally prevalent (57.1\% and 51.8\%), yet only
17.0\% of systems implement consolidation---the process by which episodic
experiences are distilled into semantic knowledge. This gap is significant:
without consolidation, episodic memories accumulate indefinitely, degrading
retrieval precision as the memory store grows. The few systems that implement
consolidation---Generative Agents' reflection generation
\cite{park2023generative}, LightMem's sleep-time processing
\cite{fang2025lightmem}, SYNAPSE's activation-based promotion
\cite{jiang2026synapse}, and CLS-inspired pipelines---draw on complementary
learning systems theory from neuroscience \cite{mcclelland1995}, suggesting
that biologically grounded approaches to consolidation are more robust than
ad-hoc summarization.

Second, the retrieval landscape shows a pronounced asymmetry: 65.2\% of systems
use vector/dense search, but only 21.4\% combine it with sparse (BM25) search,
and a mere 17.0\% employ principled fusion (typically Reciprocal Rank Fusion).
Full three-signal retrieval---sparse, dense, and graph---with fusion is
implemented by only four systems. This is notable because hybrid retrieval
consistently outperforms single-signal approaches in zero-shot settings where
labeled relevance judgments are unavailable, which is the typical case for agent
memory.

Third, lifecycle operations remain the most underserved area. Forgetting is
present in only 33.9\% of systems, and of those, the majority use simple
mechanisms (window truncation, FIFO eviction, or single-curve exponential
decay). Contradiction resolution (8.0\%) and transformation/reorganization
(12.5\%) are rarer still. The field has invested heavily in memory
\emph{formation} and \emph{retrieval} while largely neglecting memory
\emph{management}---the processes that keep a memory store healthy as it grows.

%% ---------------------------------------------------------------
\subsection{Temporal Trends}
\label{sec:temporal-trends}

Table~\ref{tab:temporal-trends} groups the surveyed systems by year of initial
publication or release and tracks the adoption of four capabilities that
distinguish sophisticated memory architectures from simple retrieval stores:
graph structure, forgetting, consolidation, and zero-dependency deployment.

\begin{table}[!t]
\centering
\caption{Temporal trends in memory system capabilities by year of publication or initial release. Percentages are relative to the total systems released in that year.}
\label{tab:temporal-trends}
\small
\begin{tabular}{@{}lrrrrrr@{}}
\toprule
& & \multicolumn{4}{c}{\textbf{\% of systems with capability}} \\
\cmidrule(l){3-6}
\textbf{Year} & \textbf{$N$} & \textbf{Graph} & \textbf{Forgetting} & \textbf{Consolidation} & \textbf{0-dep} \\
\midrule
2023 &  9 & 11.1 & 33.3 & 22.2 & 33.3 \\
2024 & 15 & 26.7 & 26.7 & 13.3 & 26.7 \\
2025 & 47 & 34.0 & 38.3 & 19.1 & 31.9 \\
2026 & 37 & 40.5 & 43.2 & 21.6 & 29.7 \\
\bottomrule
\end{tabular}
\vspace{0.3em}
\begin{flushleft}
\footnotesize
\textit{Note:} ``0-dep'' = requires zero external services (no database servers, no Docker, no cloud APIs). Year is determined by first public release or arXiv preprint date.
\end{flushleft}
\end{table}

Table~\ref{tab:temporal-trends} reveals two countervailing trends. Graph
structure has seen steady adoption, rising from 11.1\% in 2023 to 39.1\% in
2026---a trajectory consistent with the ``graph memory is table stakes''
consensus that emerged in early 2026. Similarly, forgetting mechanisms have
grown from 33.3\% to 47.8\%, driven both by Ebbinghaus-curve implementations in
production systems and by RL-learned decay policies in academic work.

In contrast, consolidation (episodic $\to$ semantic promotion) and
zero-dependency deployment have remained effectively flat across all four
years. Consolidation hovers near 20\%, suggesting that this capability
requires deep architectural commitment that most systems---especially
frameworks and coding agent tools---do not prioritize. Zero-dependency
deployment has actually declined slightly in relative terms, as the
2025--2026 wave of production systems (Mem0, Supermemory, TiMem, Papr
Memory) trend toward multi-service architectures with Neo4j, Redis, and
managed vector databases.

Figure~\ref{fig:temporal} visualizes these trends, showing the steady
rise in graph structure and forgetting adoption against the flat
trajectory of consolidation and zero-dependency deployment.

\begin{figure}[!t]
\centering
\begin{tikzpicture}
\begin{axis}[
    ybar,
    bar width=5pt,
    width=0.95\columnwidth,
    height=5.5cm,
    ylabel={\% of systems},
    symbolic x coords={2023, 2024, 2025, 2026},
    xtick=data,
    ymin=0, ymax=60,
    ytick={0,10,20,30,40,50,60},
    legend style={at={(0.5,1.08)}, anchor=south, legend columns=2,
                  font=\scriptsize},
    x tick label style={font=\small},
    y tick label style={font=\small},
    ylabel style={font=\small},
]
\addplot[fill=blue!50] coordinates
    {(2023, 11.1) (2024, 26.7) (2025, 34.0) (2026, 40.5)};
\addplot[fill=red!50] coordinates
    {(2023, 33.3) (2024, 26.7) (2025, 38.3) (2026, 43.2)};
\addplot[fill=green!50] coordinates
    {(2023, 22.2) (2024, 13.3) (2025, 19.1) (2026, 21.6)};
\addplot[fill=orange!50] coordinates
    {(2023, 33.3) (2024, 26.7) (2025, 31.9) (2026, 29.7)};
\legend{Graph, Forgetting, Consolid., 0-dep}
\end{axis}
\end{tikzpicture}
\caption{Capability adoption trends 2023--2026.  Graph structure and
forgetting show clear upward trajectories, while consolidation and
zero-dependency deployment remain flat.  Data from
Table~\ref{tab:temporal-trends}.}
\label{fig:temporal}
\end{figure}

The 2025--2026 period also marks the emergence of RL-trained memory
policies as a distinct paradigm. Memory-R1, MemRL, AgeMem, and
Mem-alpha~\cite{wang2025memalpha} all appeared in this window, collectively
demonstrating that reinforcement learning can discover effective memory
management strategies without hand-crafted heuristics. This represents the
most significant methodological shift in the field since the seminal
Generative Agents work \cite{park2023generative} established the
recency-importance-relevance retrieval formula in 2023.

%% ---------------------------------------------------------------
\subsection{Retrieval Architecture Distribution}
\label{sec:rag-distribution}

Table~\ref{tab:rag-distribution} maps the surveyed systems to the three-tier
RAG taxonomy of Gao et al.~\cite{gao2023rag}: Naive RAG (single retrieval
signal, no fusion), Advanced RAG (multiple signals or sophisticated
scoring, but fixed pipeline), and Modular RAG (composable retrieval
components with principled fusion and post-retrieval processing).  The
expansion of the coding agent category from 9 to 17 classified systems
in early 2026---driven by OpenClaw alternatives (OpenFang, IronClaw,
Nanobot) and IDE-integrated memory (Copilot, Cursor, Windsurf,
Augment Code)---has
shifted the balance toward Advanced RAG in that category, as newer
systems implement hybrid retrieval rather than simple MEMORY.md
injection.

\begin{table}[!t]
\centering
\caption{Distribution of systems across the RAG taxonomy \cite{gao2023rag}. Classification is based on the retrieval pipeline as documented; systems that delegate retrieval entirely to an LLM (e.g., Letta's agent-driven tool calls) are classified as Naive with an agent-augmented qualifier.}
\label{tab:rag-distribution}
\small
\begin{tabular}{@{}lrrp{4.0cm}@{}}
\toprule
\textbf{RAG Type} & \textbf{Count} & \textbf{\%} & \textbf{Representative Systems} \\
\midrule
Naive RAG       & 48 & 42.9 & LangChain, LlamaIndex, Engram, Agency Swarm \\
Advanced RAG    & 44 & 39.3 & Mem0, Generative Agents, Hindsight, A-MEM \\
Modular RAG     & 20 & 17.9 & Zep/Graphiti, SYNAPSE, HippoRAG, QMD \\
\bottomrule
\end{tabular}
\end{table}

A plurality (42.9\%) of memory systems use Naive RAG---typically a single
vector similarity query or direct context injection with no fusion or
reranking. This is especially common in framework memory modules (LangChain,
LlamaIndex, Haystack, Agency Swarm) and coding agent tools (Engram,
MemoryMesh, ClaudeHistory Cloud), where memory is a secondary concern appended
to a larger orchestration framework. Advanced RAG systems (39.3\%) add
multi-signal scoring (e.g., Generative Agents' recency $\times$ importance
$\times$ relevance formula) or sophisticated single-signal pipelines (e.g.,
Memvid's Tantivy FTS + HNSW with chronological reranking). Modular RAG
(17.9\%) requires composable, independently configurable retrieval components
with principled fusion---a design pattern found almost exclusively in dedicated
memory systems and research architectures that treat retrieval as a first-class
architectural concern.

%% ---------------------------------------------------------------
\subsection{Benchmark Cross-Comparison}
\label{sec:benchmark-cross-comparison}

\input{sections/benchmark-table-fragment}

%% ---------------------------------------------------------------
\subsubsection{Baseline Replication}
\label{sec:baseline-replication}

To provide a controlled comparison with full methodological transparency, we
replicate the two canonical baselines---full-context injection and na\"ive
vector RAG---on both LoCoMo and LongMemEval\_S using identical infrastructure.
The generator is Gemini-2.0-Flash-001 via OpenRouter (temperature~0,
max\_tokens~512); the judge is GPT-4o-mini via OpenRouter (temperature~0,
max\_tokens~10), using the binary yes/no LLM-as-Judge protocol from
\cite{maharana2024locomo}.  For LoCoMo, we evaluate 1{,}540 questions across
categories~1--4 (single-hop, temporal, multi-hop, open-domain), excluding
category~5 (adversarial) per the standard evaluation protocol.  For
LongMemEval\_S, we evaluate all 500 questions across six question types.
Na\"ive RAG uses ChromaDB with \texttt{all-MiniLM-L6-v2} embeddings and
top-10 cosine similarity retrieval.  Retry policy: 5~attempts with
exponential backoff (1/2/4/8/16\,s).

\smallskip
\noindent\textbf{Statistical methods.}\enspace
Confidence intervals use the Wilson score interval~\cite{wilson1927} for
binomial proportions, which provides correct coverage near boundary values
unlike the normal approximation.  Significance testing uses McNemar's
test~\cite{mcnemar1947} with continuity correction on the $2 \times 2$
contingency table of per-question agreement between systems---the appropriate
test for paired nominal data where each question yields a binary outcome from
both systems.  Effect sizes are reported as differences in proportions with
95\% confidence intervals.

Table~\ref{tab:our-baselines} and Figure~\ref{fig:crossover} report overall
results; Tables~\ref{tab:locomo-categories} and~\ref{tab:longmemeval-types}
decompose accuracy by question category and type;
Table~\ref{tab:mcnemar} reports the McNemar contingency tables and test
statistics; and Figure~\ref{fig:breakdown} visualizes the per-category and
per-type breakdowns.

\begin{table}[!t]
\centering
\caption{Baseline replication with Gemini-2.0-Flash-001 (generator) and
GPT-4o-mini (judge).  95\% confidence intervals via Wilson score interval.
McNemar's test confirms both differences are statistically significant
($p < 0.001$).  Full results and code at
\url{https://github.com/SecurityRonin/alaya/tree/main/bench}.}
\label{tab:our-baselines}
\small
\begin{tabular}{@{}lcccc@{}}
\toprule
& \multicolumn{2}{c}{\textbf{LoCoMo}} & \multicolumn{2}{c}{\textbf{LongMemEval}} \\
\cmidrule(lr){2-3} \cmidrule(lr){4-5}
\textbf{System} & \textbf{Acc.} & \textbf{95\% CI} & \textbf{Acc.} & \textbf{95\% CI} \\
\midrule
Full-context & 61.4 & [58.9, 63.8] & 46.2 & [41.9, 50.6] \\
Na\"ive RAG  & 26.0 & [23.9, 28.2] & \textbf{54.6} & [50.2, 58.9] \\
\midrule
$\Delta$ (FC $-$ RAG) & $+35.4$ & & $-8.4$ & \\
\bottomrule
\end{tabular}
\end{table}

\begin{figure}[!t]
\centering
\begin{tikzpicture}
\begin{axis}[
    ybar,
    bar width=14pt,
    width=0.95\columnwidth,
    height=5.5cm,
    ylabel={Accuracy (\%)},
    symbolic x coords={LoCoMo, LongMemEval\_S},
    xtick=data,
    ymin=0, ymax=80,
    ytick={0,20,40,60,80},
    legend style={at={(0.5,1.05)}, anchor=south, legend columns=2,
                  font=\small},
    nodes near coords,
    nodes near coords style={font=\scriptsize},
    every node near coord/.append style={yshift=2pt},
]
\addplot[fill=blue!50] coordinates
    {(LoCoMo, 61.4) (LongMemEval\_S, 46.2)};
\addplot[fill=red!50] coordinates
    {(LoCoMo, 26.0) (LongMemEval\_S, 54.6)};
\legend{Full-context, Na\"ive RAG}
\end{axis}
\end{tikzpicture}
\caption{The retrieval crossover: full-context injection dominates on
LoCoMo (${\sim}$16--26K tokens) but na\"ive RAG outperforms on
LongMemEval (${\sim}$115K tokens).  McNemar's test confirms both
differences are significant ($p < 0.001$).}
\label{fig:crossover}
\end{figure}

\begin{table}[!t]
\centering
\caption{McNemar's test on paired per-question outcomes.  Each cell counts
questions where systems agree or disagree.  The extreme asymmetry
(FC-only $\gg$ RAG-only on LoCoMo; RAG-only $>$ FC-only on LongMemEval)
drives the significant $\chi^2$ values.  $\Delta$~CI is the 95\%
confidence interval for the difference in proportions.}
\label{tab:mcnemar}
\small
\footnotesize
\begin{tabular}{@{}lrrrr@{}}
\toprule
& \multicolumn{2}{c}{\textbf{LoCoMo} ($n\!=\!1529$\textsuperscript{*})}
& \multicolumn{2}{c}{\textbf{LongMemEval} ($n\!=\!500$)} \\
\cmidrule(lr){2-3} \cmidrule(lr){4-5}
& RAG $\checkmark$ & RAG $\times$ & RAG $\checkmark$ & RAG $\times$ \\
\midrule
FC $\checkmark$ & 353 & 583 & 180 & 51 \\
FC $\times$     &  44 & 549 &  93 & 176 \\
\midrule
\multicolumn{1}{@{}l}{$\chi^2$ (corrected)}
  & \multicolumn{2}{c}{461.6}
  & \multicolumn{2}{c}{11.7} \\
\multicolumn{1}{@{}l}{$p$-value}
  & \multicolumn{2}{c}{$< 10^{-5}$}
  & \multicolumn{2}{c}{$6.3 \times 10^{-4}$} \\
\multicolumn{1}{@{}l}{$\Delta$ (FC $-$ RAG)}
  & \multicolumn{2}{c}{$+35.4$~pp}
  & \multicolumn{2}{c}{$-8.4$~pp} \\
\multicolumn{1}{@{}l}{95\% CI of $\Delta$}
  & \multicolumn{2}{c}{$[+32.1, +38.7]$}
  & \multicolumn{2}{c}{$[-14.5, -2.3]$} \\
\bottomrule
\end{tabular}
\vspace{0.2em}
\begin{flushleft}
\footnotesize
\textsuperscript{*}11 duplicate question texts removed for pairing.
\end{flushleft}
\end{table}

Three findings merit discussion.

\smallskip
\noindent\textbf{Finding 1: The retrieval crossover is statistically
significant in both directions.}\enspace
On LoCoMo, full-context outperforms na\"ive RAG by 35.4~pp
(McNemar's $\chi^2 = 461.6$, $p < 10^{-5}$, $n = 1{,}529$ paired questions;
effect size 95\% CI: $[+32.1, +38.7]$~pp);
on LongMemEval, na\"ive RAG outperforms full-context by 8.4~pp
($\chi^2 = 11.7$, $p = 6.3 \times 10^{-4}$, $n = 500$;
effect size 95\% CI: $[-14.5, -2.3]$~pp).
Table~\ref{tab:mcnemar} shows the full contingency tables: the LoCoMo
asymmetry is extreme (583~questions correct for FC only vs.\ 44 for RAG
only), while the LongMemEval asymmetry is moderate (93 vs.\ 51) but
statistically significant.  The crossover
occurs because LoCoMo conversations average 16--26K tokens---well within the
generator's context window, allowing full-context injection to preserve all
temporal and relational cues that top-10 retrieval discards.  LongMemEval
histories average ${\sim}$115K tokens, exceeding the model's effective
attention span; injecting everything forces the model to locate relevant
details among overwhelming noise, while RAG's filtering---even crude cosine
similarity with a fixed top-$k$---improves signal-to-noise ratio.

Note that our full-context LoCoMo accuracy (61.4\%) is lower than the GPT-4o
baseline (${\sim}$72.9\%) in Table~\ref{tab:benchmark-comparison}, reflecting
the model quality gap between Gemini-2.0-Flash and GPT-4o rather than a
methodological difference.  The crossover pattern is architecture-dependent,
not model-dependent: within any fixed model, the relative advantage of
full-context versus RAG reverses as history length exceeds the model's
effective attention window.

\smallskip
\noindent\textbf{Finding 2: Category analysis reveals systematic weaknesses
in both baselines.}\enspace
Tables~\ref{tab:locomo-categories} and~\ref{tab:longmemeval-types} decompose
accuracy by question category and type, with the visual comparison in
Figure~\ref{fig:breakdown}.

On LoCoMo, temporal reasoning (category~2) is catastrophically poor for both
systems: 19.3\% for full-context and 6.5\% for na\"ive RAG---well below
random guessing on binary questions.  These questions require reasoning about
\emph{when} events occurred relative to each other---a capability that
neither context stuffing nor cosine similarity provides.  Multi-hop questions
(category~3) show a similar pattern at higher absolute levels (46.9\% vs.\
21.9\%).  Open-domain questions (category~4), which comprise 54.6\% of all
LoCoMo questions and reward broad coverage rather than precise relational
reasoning, reach the highest accuracy (81.5\% full-context), inflating the
overall score.  Excluding category~4, full-context accuracy drops to 37.2\%.

On LongMemEval (Table~\ref{tab:longmemeval-types}), single-session user
queries show the largest RAG advantage, jumping from 60.0\% (FC) to 88.6\%
(RAG)---facts stated by the user in a single session are semantically close
to the question and well-served by embedding similarity.  Single-session
preference questions are poor for both systems (10.0\% FC, 33.3\% RAG);
preferences are often implied rather than stated explicitly, making them
difficult for both approaches.  Multi-session queries improve from 30.8\% to
44.4\%, benefiting from noise filtering.  But temporal reasoning is
\emph{identical} at 29.3\% for both systems, confirming the LoCoMo finding:
temporal reasoning requires structured temporal indexing, not content
similarity.  Single-session assistant queries are near-ceiling for both
(96.4\% FC, 94.6\% RAG)---facts stated by the assistant are typically clear,
structured responses that both approaches handle well.

\begin{table}[!t]
\centering
\caption{LoCoMo accuracy by question category. Category~5 (adversarial)
excluded per evaluation protocol.  The temporal reasoning gap
(category~2) is the most pronounced weakness for both baselines.}
\label{tab:locomo-categories}
\small
\begin{tabular}{@{}clrrrr@{}}
\toprule
\textbf{Cat.} & \textbf{Type} & \textbf{$N$} & \textbf{FC (\%)} & \textbf{RAG (\%)} & \textbf{$\Delta$} \\
\midrule
1 & Single-hop     & 282 & 54.3 & 13.5 & $+40.8$ \\
2 & Temporal        & 321 & 19.3 &  6.5 & $+12.8$ \\
3 & Multi-hop       &  96 & 46.9 & 21.9 & $+25.0$ \\
4 & Open-domain     & 841 & 81.5 & 38.0 & $+43.5$ \\
\midrule
  & \textbf{Overall} & 1{,}540 & 61.4 & 26.0 & $+35.4$ \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[!t]
\centering
\caption{LongMemEval accuracy by question type.  Single-session user queries
show the largest RAG advantage ($+28.6$~pp); temporal reasoning is identical
for both systems (29.3\%), confirming the LoCoMo finding.  Single-session
preference questions are poor for both approaches.}
\label{tab:longmemeval-types}
\small
\begin{tabular}{@{}lrrrr@{}}
\toprule
\textbf{Type} & \textbf{$N$} & \textbf{FC (\%)} & \textbf{RAG (\%)} & \textbf{$\Delta$} \\
\midrule
SS-user       &  70 & 60.0 & 88.6 & $-28.6$ \\
SS-assistant  &  56 & 96.4 & 94.6 & $+1.8$ \\
SS-preference &  30 & 10.0 & 33.3 & $-23.3$ \\
Multi-session & 133 & 30.8 & 44.4 & $-13.5$ \\
Temporal      & 133 & 29.3 & 29.3 & $0.0$ \\
Knowledge update & 78 & 66.7 & 64.1 & $+2.6$ \\
\midrule
\textbf{Overall} & 500 & 46.2 & 54.6 & $-8.4$ \\
\bottomrule
\end{tabular}
\end{table}

\begin{figure*}[!t]
\centering
\begin{minipage}[t]{0.48\textwidth}
\centering
\begin{tikzpicture}[baseline=(current axis.north)]
\begin{axis}[
    ybar,
    bar width=8pt,
    width=\textwidth,
    height=5.5cm,
    ylabel={Accuracy (\%)},
    symbolic x coords={Single-hop, Temporal, Multi-hop, Open-domain},
    xtick=data,
    x tick label style={rotate=25, anchor=east, font=\scriptsize},
    ymin=0, ymax=110,
    ytick={0,20,40,60,80,100},
    legend style={at={(0.5,1.08)}, anchor=south, legend columns=2,
                  font=\scriptsize},
    nodes near coords,
    nodes near coords style={font=\tiny},
    every node near coord/.append style={yshift=2pt},
    title style={yshift=3pt},
    title={\small (a) LoCoMo by category},
    clip=false,
]
\addplot[fill=blue!50] coordinates
    {(Single-hop, 54.3) (Temporal, 19.3) (Multi-hop, 46.9) (Open-domain, 81.5)};
\addplot[fill=red!50] coordinates
    {(Single-hop, 13.5) (Temporal, 6.5) (Multi-hop, 21.9) (Open-domain, 38.0)};
\legend{Full-context, Na\"ive RAG}
\end{axis}
\end{tikzpicture}
\end{minipage}\hfill
\begin{minipage}[t]{0.48\textwidth}
\centering
\begin{tikzpicture}[baseline=(current axis.north)]
\begin{axis}[
    ybar,
    bar width=6pt,
    width=\textwidth,
    height=5.5cm,
    ylabel={Accuracy (\%)},
    symbolic x coords={SS-User, SS-Asst, SS-Pref, Multi-S, Temporal, K-Upd},
    xtick=data,
    x tick label style={rotate=25, anchor=east, font=\scriptsize},
    ymin=0, ymax=110,
    ytick={0,20,40,60,80,100},
    legend style={at={(0.5,1.08)}, anchor=south, legend columns=2,
                  font=\scriptsize},
    title style={yshift=3pt},
    title={\small (b) LongMemEval by type},
    clip=false,
]
\addplot[fill=blue!50, nodes near coords, nodes near coords style={font=\tiny, yshift=5pt}] coordinates
    {(SS-User, 60.0) (SS-Asst, 96.4) (SS-Pref, 10.0) (Multi-S, 30.8) (Temporal, 29.3) (K-Upd, 66.7)};
\addplot[fill=red!50, nodes near coords, nodes near coords style={font=\tiny, yshift=-1pt}] coordinates
    {(SS-User, 88.6) (SS-Asst, 94.6) (SS-Pref, 33.3) (Multi-S, 44.4) (Temporal, 29.3) (K-Upd, 64.1)};
\legend{Full-context, Na\"ive RAG}
\end{axis}
\end{tikzpicture}
\end{minipage}
\caption{Per-category and per-type accuracy breakdown.  (a)~LoCoMo: temporal
reasoning (cat.~2) is catastrophic for both systems; open-domain (cat.~4)
inflates overall full-context accuracy.  (b)~LongMemEval: single-session user
queries show the largest RAG advantage; temporal reasoning is identical at
29.3\% for both systems.  Error bars omitted for clarity; confidence intervals
are in Tables~\ref{tab:locomo-categories} and~\ref{tab:longmemeval-types}.}
\label{fig:breakdown}
\end{figure*}

\smallskip
\noindent\textbf{Finding 3: Neither baseline addresses what lifecycle
management is designed for.}\enspace
The crossover demonstrates that \emph{how much} context to include is at
least as important as \emph{what} to include---yet neither existing benchmark
evaluates adaptive retrieval scope.  Memory systems with lifecycle
management (consolidation, forgetting, graph-based association) are designed
precisely for the regime where neither extreme is optimal: they reduce noise
through forgetting, improve signal through consolidation, and discover
indirect connections through graph traversal.  These capabilities should
manifest as performance that exceeds both baselines simultaneously on both
benchmarks, but no published evaluation tests this hypothesis.  This gap
reinforces the evaluation methodology problem discussed in
\S\ref{sec:open-problems}: the field needs benchmarks that test memory
\emph{management}, not just memory \emph{retrieval}.

%% ---------------------------------------------------------------
\subsubsection{The Emerging Benchmark Landscape}
\label{sec:benchmark-landscape}

The field's reliance on LoCoMo, LongMemEval, and DMR as the primary
evaluation benchmarks is increasingly insufficient. Three recent benchmarks
address distinct gaps that these established benchmarks leave open.

\textbf{TReMu}~\cite{ge2025tremu} targets temporal reasoning in
multi-session dialogues, constructing multiple-choice questions from
LoCoMo conversations that require temporal anchoring, precedence, and
interval reasoning.  GPT-4o achieves only 29.8\% via standard prompting
(77.7\% with the authors' neuro-symbolic framework), confirming that
temporal reasoning---the weakest category in both our LoCoMo and
LongMemEval baselines (Table~\ref{tab:locomo-categories},
Table~\ref{tab:longmemeval-types})---requires structured temporal
indexing rather than content similarity.

\textbf{LoCoMo-Plus}~\cite{li2026locomoplus} extends LoCoMo beyond
factual recall to \emph{cognitive memory}: questions whose correct
answers depend on implicit constraints (causal, state, goal, value)
rather than explicitly stated facts.  This cue--trigger semantic
disconnect reveals failures invisible to factual benchmarks, as models
must retain and apply latent constraints across long contexts.

\textbf{MemoryAgentBench} (MAB)~\cite{hu2025mab} provides the most
comprehensive coverage, evaluating four core competencies via
incremental multi-turn interactions: \emph{accurate retrieval},
\emph{test-time learning}, \emph{long-range understanding}, and
\emph{conflict resolution} (selective forgetting).  Crucially, MAB is
the first benchmark to explicitly test forgetting quality and
contradiction handling---capabilities that our feature prevalence
analysis (\S\ref{sec:feature-prevalence}) identifies as the most
underserved in the field.

Table~\ref{tab:benchmark-landscape} summarizes how these six benchmarks
collectively map to the lifecycle operations from our taxonomy.  The
progression from LoCoMo (factual retrieval) through TReMu (temporal
reasoning) to MAB (lifecycle competencies) mirrors the field's gradual
recognition that memory evaluation must encompass management processes,
not just retrieval accuracy.

\begin{table}[!t]
\centering
\caption{Benchmark landscape for agent memory evaluation.  Columns
indicate which memory competencies each benchmark tests.
$\bullet$~=~primary focus; $\circ$~=~partial coverage.
Only MAB explicitly evaluates forgetting and contradiction resolution.}
\label{tab:benchmark-landscape}
\scriptsize
\setlength{\tabcolsep}{3pt}
\begin{tabular}{@{}lcccccc@{}}
\toprule
& \rotatebox{70}{\textbf{Factual}} & \rotatebox{70}{\textbf{Temporal}} & \rotatebox{70}{\textbf{Cognitive}} & \rotatebox{70}{\textbf{Learning}} & \rotatebox{70}{\textbf{Forgetting}} & \rotatebox{70}{\textbf{Understand.}} \\
\midrule
LoCoMo         & $\bullet$ & $\circ$ &           &           &           &           \\
LongMemEval    & $\bullet$ & $\circ$ &           &           &           &           \\
DMR            & $\bullet$ &         &           &           &           &           \\
TReMu          &           & $\bullet$ &         &           &           &           \\
LoCoMo-Plus    & $\bullet$ &         & $\bullet$ &           &           &           \\
MAB            & $\bullet$ &         &           & $\bullet$ & $\bullet$ & $\bullet$ \\
\bottomrule
\end{tabular}
\end{table}

%% ---------------------------------------------------------------
\subsubsection{MAB Baseline Replication}
\label{sec:mab-replication}

To validate our baselines on a benchmark that tests lifecycle competencies
rather than retrieval alone, we replicate the full-context and na\"ive RAG
baselines on MemoryAgentBench using the same infrastructure as
\S\ref{sec:baseline-replication}: Gemini-2.0-Flash-001 (generator,
temperature~0, max\_tokens~512) and GPT-4o-mini (judge, temperature~0,
max\_tokens~10).  We evaluate all four MAB competencies: accurate retrieval
(AR, 500~questions from EventQA), test-time learning (TTL, 100~questions
from ICL-NLU), long-range understanding (LRU, 34~questions from
DetectiveQA), and conflict resolution (CR, 100~questions from
FactConsolidation).  Table~\ref{tab:mab-baselines} reports the results.

\begin{table}[!t]
\centering
\caption{MAB baseline replication with Gemini-2.0-Flash-001 (generator)
and GPT-4o-mini (judge).  AR = accurate retrieval, TTL = test-time
learning, LRU = long-range understanding, CR = conflict resolution
(selective forgetting).  The TTL gap (42.0~pp) confirms that in-context
learning requires full conversation history; the CR results confirm that
neither baseline handles contradictions effectively.}
\label{tab:mab-baselines}
\footnotesize
\begin{tabular}{@{}lrrrr@{}}
\toprule
\textbf{Competency} & \textbf{$N$} & \textbf{FC (\%)} & \textbf{RAG (\%)} & \textbf{$\Delta$} \\
\midrule
AR (Accurate Retrieval)     & 500 & 94.0 & 90.4 & $+3.6$ \\
TTL (Test-Time Learning)    & 100 & 86.0 & 44.0 & $+42.0$ \\
LRU (Long-Range Underst.)   &  34 & 82.4 & 67.6 & $+14.7$ \\
CR (Conflict Resolution)    & 100 & 50.0 & 41.0 & $+9.0$ \\
\bottomrule
\end{tabular}
\end{table}

\begin{figure}[!t]
\centering
\begin{tikzpicture}
\begin{axis}[
    ybar,
    bar width=10pt,
    width=0.95\columnwidth,
    height=5.5cm,
    ylabel={Accuracy (\%)},
    symbolic x coords={AR, TTL, LRU, CR},
    xtick=data,
    ymin=0, ymax=110,
    ytick={0,20,40,60,80,100},
    legend style={at={(0.5,1.05)}, anchor=south, legend columns=2,
                  font=\small},
    nodes near coords,
    nodes near coords style={font=\scriptsize},
    every node near coord/.append style={yshift=2pt},
]
\addplot[fill=blue!50] coordinates
    {(AR, 94.0) (TTL, 86.0) (LRU, 82.4) (CR, 50.0)};
\addplot[fill=red!50] coordinates
    {(AR, 90.4) (TTL, 44.0) (LRU, 67.6) (CR, 41.0)};
\legend{Full-context, Na\"ive RAG}
\end{axis}
\end{tikzpicture}
\caption{MAB baseline accuracy by competency.  Test-time learning (TTL)
shows the largest gap ($+42.0$~pp): RAG destroys the sequential structure
needed for in-context learning.  Conflict resolution (CR) is near chance
for both baselines, confirming that contradiction handling requires
explicit lifecycle management.}
\label{fig:mab-baselines}
\end{figure}

\smallskip
\noindent\textbf{Finding 4: Lifecycle competencies expose systematic
failures invisible to retrieval-only benchmarks.}\enspace
The MAB results reveal a capability hierarchy that LoCoMo and LongMemEval
cannot distinguish.  Accurate retrieval (AR) is nearly solved by both
baselines---94.0\% full-context, 90.4\% na\"ive RAG---confirming that
factual lookup over moderate-length contexts is no longer a discriminating
test.  Test-time learning (TTL) produces the largest gap in our entire
evaluation: 86.0\% vs.\ 44.0\% ($\Delta = +42.0$~pp).  TTL questions
require the model to learn task patterns from in-context examples; RAG's
top-$k$ retrieval selects semantically similar chunks but discards the
sequential structure that makes in-context learning possible.  This echoes
the LoCoMo temporal reasoning finding (Finding~2): capabilities that
depend on \emph{structural} rather than \emph{semantic} properties of the
conversation history are systematically destroyed by na\"ive retrieval.

Long-range understanding (LRU) shows a moderate gap (82.4\% vs.\ 67.6\%),
indicating that cross-document reasoning benefits from full context but is
partially recoverable through retrieval when the relevant passages share
vocabulary with the question.  Most strikingly, conflict resolution (CR)
is near chance for both baselines (50.0\% vs.\ 41.0\%), confirming that
\emph{neither} full-context injection \emph{nor} retrieval provides a
mechanism for resolving contradictions between earlier and later
information.  This result directly validates the gap identified in
Table~\ref{tab:feature-prevalence}: only 8.0\% of surveyed systems
implement contradiction resolution, and the CR baseline results show why
this capability cannot emerge from retrieval alone---it requires explicit
lifecycle management that identifies and resolves conflicting facts.

%% ---------------------------------------------------------------
\subsection{Function-Based vs.\ Type-Based Memory Organization}
\label{sec:function-vs-type}

The CoALA taxonomy~\cite{sumers2024coala} organizes memory by
\emph{type}---episodic, semantic, procedural---and this typology
underpins our classification framework.  However, several systems have
independently converged on an orthogonal organizing principle:
decomposing memory by \emph{function} (what purpose it serves) rather
than type (what kind of information it stores).

ReMe~\cite{reme2025} distinguishes \emph{personal memory} (user
preferences and traits) from \emph{task memory} (learned strategies from
prior executions), treating these as first-class architectural
boundaries rather than metadata annotations on a unified store.
GitHub Copilot Memory~\cite{copilot2026} similarly separates memory
by function, with citation-linked code memories serving a different
retrieval and validation pipeline than general user preferences.
Augment Code~\cite{augment2026} implicitly adopts a functional split
between its real-time Context Engine (codebase-structural memory) and
its Memories feature (cross-session knowledge).

This functional decomposition suggests the field may be converging on a
two-dimensional classification: \emph{type} (what kind of memory)
$\times$ \emph{function} (what purpose it serves).  A fact about a
user's coding style is semantically a ``semantic memory'' in CoALA
terms, but functionally a ``personal memory'' in ReMe's taxonomy---and
the appropriate retrieval strategy, update policy, and lifecycle
management may differ based on function rather than type.  Whether
function-based decomposition produces measurably better retrieval or
lifecycle management than type-based organization remains an open
empirical question, but the independent convergence across three
unrelated systems suggests that the CoALA typology, while foundational,
may be insufficient as the sole organizational axis for agent memory.

%% ---------------------------------------------------------------
\subsection{The Long Tail Problem}
\label{sec:long-tail}

The feature prevalence data in Table~\ref{tab:feature-prevalence} reveals a
pronounced long-tail distribution: most capabilities are concentrated in a
small number of systems, while the majority of the 116 surveyed systems
implement only basic memory functionality. To quantify this, we define a
\emph{feature breadth count} as the number of features present from a set of
eight discriminating capabilities: episodic memory, semantic memory, graph
structure, forgetting, consolidation, contradiction resolution, preference
learning, and hybrid retrieval fusion.  This count measures architectural
breadth, not quality or depth---a system excelling at graph-based retrieval
may have a low count while outperforming broader systems on specific tasks.
Table~\ref{tab:capability-distribution} shows the distribution.

\begin{table}[!t]
\centering
\caption{Distribution of feature breadth counts across 116 systems.  The count tallies eight features: episodic memory, semantic memory, graph structure, forgetting, consolidation, contradiction resolution, preference learning, and hybrid retrieval.  This measures architectural breadth, not quality.}
\label{tab:capability-distribution}
\small
\begin{tabular}{@{}crrp{3.8cm}@{}}
\toprule
\textbf{Count} & \textbf{Sys.} & \textbf{\%} & \textbf{Examples} \\
\midrule
0--1 & 28 & 25.0 & Agency Swarm, Qwen-Agent, AutoGPT, Nanobot \\
2--3 & 40 & 35.7 & LangChain, Engram, Memvid, HippoRAG, CAMEL \\
4--5 & 25 & 22.3 & Mem0, Hindsight, SYNAPSE, LightMem, memU \\
6--7 & 12 & 10.7 & Zep/Graphiti, EverMemOS, MemOS, Vestige, ReMe \\
8    &  2 &  1.8 & --- \\
\bottomrule
\end{tabular}
\end{table}

The distribution is sharply skewed: 60.7\% of systems implement three or
fewer of the eight capabilities. Only 12.5\% implement six or more. This concentration has three implications for the
field.

First, the median memory system is a \textbf{vector store with optional
context windowing}---it stores text, retrieves by embedding similarity, and
manages capacity through truncation or FIFO eviction. The CoALA framework
\cite{sumers2024coala} classifies this as working memory with optional
episodic storage, the simplest configuration in its taxonomy. The gap between
what the literature proposes (multi-store architectures with lifecycle
processes) and what practitioners ship (single-store vector retrieval) remains
substantial.

Second, capabilities cluster rather than distribute uniformly. Systems that
implement consolidation almost always also implement forgetting (17 of 19, or
89.5\%), but the reverse is not true---many systems with forgetting (38 total)
lack consolidation. Similarly, graph structure strongly predicts hybrid
retrieval fusion: 16 of 19 systems with fusion also have graph traversal.
These correlations suggest that certain capabilities are architecturally
co-dependent: consolidation requires a semantic store to consolidate
\emph{into}, which in turn motivates forgetting to manage the growing semantic
store; graph traversal provides a third retrieval signal that motivates
principled fusion to combine it with sparse and dense signals.

Third, the long tail creates an \textbf{evaluation gap}. Benchmarks like
LoCoMo and LongMemEval evaluate memory recall and factual accuracy, but they
do not test forgetting quality, preference tracking, contradiction resolution,
or associative retrieval through graph structure. As a result, simple systems
that score well on retrieval benchmarks may be indistinguishable from
architecturally richer systems in published evaluations, obscuring the
practical benefits of lifecycle operations that only manifest over extended
use. The field needs benchmarks that evaluate memory \emph{management}, not
just memory \emph{retrieval}---a point we return to in
\S\ref{sec:open-problems}.

\input{sections/system-matrix}
