\section{Background}
\label{sec:background}

This section presents the neuroscience and Buddhist psychology foundations
that inform \alaya{}'s design, with particular attention to their
convergence on the same structural principles for memory architecture.

\subsection{Complementary Learning Systems}

McClelland~\etal~\cite{mcclelland1995} proposed that biological memory
relies on two complementary systems: a fast-learning hippocampal system that
encodes episodes with high fidelity, and a slow-learning neocortical system
that gradually extracts stable patterns through interleaved replay. This
separation prevents catastrophic interference---new learning destroying old
knowledge---while enabling generalization from accumulated experience.

\alaya{} instantiates this with an episodic store (rapid, high-fidelity
conversation recording) and a semantic store (stable, deduplicated knowledge
extracted during consolidation). The consolidation operation implements
CLS-style interleaved replay: batches of recent episodes are processed by an
LLM provider to extract facts, relationships, events, and concepts, which
are stored as semantic nodes distinct in kind from their source episodes.

\subsection{Bjork Dual-Strength Forgetting}

Bjork and Bjork~\cite{bjork1992} distinguished two independent strengths for
every memory: \emph{storage strength} (how deeply encoded) and
\emph{retrieval strength} (how currently accessible). Storage strength
increases monotonically with exposure; retrieval strength decays over time
but recovers dramatically when a memory is successfully retrieved---the
\emph{spacing effect}. Anderson~\etal~\cite{anderson1994} extended this with
\emph{retrieval-induced forgetting} (RIF): retrieving one memory actively
suppresses competitors sharing the same retrieval cues.

\alaya{} implements both mechanisms. Each memory node carries independent
storage and retrieval strengths. Retrieval boosts both strengths; the passage
of time decays only retrieval strength. Competitors sharing retrieval cues
are suppressed during query, implementing RIF. The \texttt{forget()} operation
uses retrieval strength to determine which memories are effectively
inaccessible.

\subsection{Hebbian Learning and Spreading Activation}

Hebb's principle~\cite{hebb1949}---``neurons that fire together wire
together''---describes use-dependent synaptic strengthening. Collins and
Loftus~\cite{collins1975} formalized spreading activation in semantic
networks: activating one concept propagates to associated concepts through
weighted links with distance-based decay.

\alaya{} maintains a Hebbian graph overlay across all memory stores. Links
are created during encoding (temporal, topical, entity, causal) and during
retrieval (co-retrieval links between jointly accessed memories). Link
weights strengthen through repeated co-activation and decay through disuse,
producing emergent small-world topology. Spreading activation is used during
retrieval to discover indirect associations---a query about ``cooking'' can
surface chemistry-related memories through shared Hebbian links.

\subsection{Yogac\=ara Process Model}
\label{sec:yogacara}

The Yogac\=ara (\cjk{瑜伽行派}) school of Buddhist philosophy,
systematized in the \emph{Yogac\=arabh\=umi-\\'s\=astra}
(\cjk{瑜伽師地論}, attributed to Maitreya/Asa\.{n}ga, translated by
Xuanzang 646--648~CE)~\cite{yogacarabhumi} and the \emph{Cheng Weishi
Lun} (\cjk{成唯識論}, Xuanzang, 659~CE)~\cite{chengweishilun},
developed a detailed phenomenology of memory more than a millennium
before computational cognitive science.

\subsubsection{The Storehouse and Seeds}

The \emph{\=alaya-vij\~n\=ana} (\cjk{阿賴耶識}, storehouse
consciousness) is the eighth and deepest consciousness in Yogac\=ara
theory---a continuous stream that underlies all mental activity.  It
receives impressions from every experience and stores them as
\emph{b\=ija} (\cjk{種子}, seeds): latent potentials that are not
static data but living dispositions with conditions for ripening.
The key verse from the \emph{Sa\d{m}dhinirmocana S\=utra}
(\cjk{解深密經})~\cite{samdhinirmocana} captures this character:

\begin{quote}
\emph{The \=ad\=ana consciousness is exceedingly deep and subtle; all
its seeds are like a torrential flood.}
\end{quote}

Seeds interact with one another: they have mutual dependence, ripen
heterogeneously (\emph{vip\=aka}, \cjk{異熟}---the result differs in
nature from the cause), and are continuously modified by new experience.
Above the storehouse, \emph{manas} (\cjk{末那識}, the seventh
consciousness) performs deliberative reflection, continuously examining
and reinterpreting what the storehouse contains---mirroring the CLS
distinction between fast episodic encoding and slow semantic
integration~\cite{mcclelland1995}.

\subsubsection{Perfuming and Transformation}

\emph{V\=asan\=a} (\cjk{薰習}, perfuming) is the process by which each
experience leaves a subtle trace on the storehouse, ``like incense
gradually permeating cloth''~\cite{chengweishilun}.  No single
experience is transformative; preferences emerge from the accumulation
of many similar traces.

The complementary process is \emph{\=a\'sraya-par\=av\d{r}tti}
(\cjk{轉依}, transformation of the basis): a periodic restructuring of
the storehouse itself---resolving contradictions, pruning stale
dispositions, and moving the substrate toward greater coherence.

\Cref{tab:yogacara-mapping} maps these concepts to \alaya{}'s
architecture.

\begin{table}[t]
\centering
\caption{Mapping Yogac\=ara Concepts to \alaya{} Architecture}
\label{tab:yogacara-mapping}
\begin{tabular}{@{}L{2.0cm}L{1.8cm}L{3.2cm}@{}}
\toprule
\textbf{Yogac\=ara} & \textbf{Axis} & \textbf{\alaya{} Operation} \\
\midrule
\emph{\=Alaya-vij\~n\=ana} & Structure & Persistent memory store \\
\emph{B\=ija} (seeds) & Structure & Memory nodes with strengths \\
\emph{V\=asan\=a} & Preference & Impression accumulation \\
\emph{Vip\=aka} & Lifecycle & Consolidation \\
\emph{\=A\'sraya-par\=av\d{r}tti} & Lifecycle & Transformation \\
\emph{Vijnapti-m\=atra} & Retrieval & Encoding specificity \\
\bottomrule
\end{tabular}
\end{table}

\noindent These principles map directly to \alaya{}'s lifecycle
operations, as detailed in \Cref{sec:architecture}.
