\section{Discussion}
\label{sec:discussion}

\subsection{Yogac\=ara as Architectural Guide}

We identify four cases where Yogac\=ara concepts predict design choices
that the AI memory community has independently converged
upon~\cite{hui2026taxonomy}. \alaya{} was designed with these predictions
as explicit architectural constraints rather than post-hoc rationalizations:

\begin{itemize}[nosep]
    \item \textbf{Heterogeneous ripening} (\emph{vip\=aka}) predicted that
    consolidation should produce outputs qualitatively different from inputs.
    \alaya{}'s consolidation converts time-bound episodes into
    decontextualized, typed semantic nodes---a structural transformation,
    not a summary.
    \item \textbf{Gradual perfuming} (\emph{v\=asan\=a}) predicted that
    preferences should emerge from accumulation, not extraction. \alaya{}'s
    five-impression threshold before preference crystallization implements
    this directly.
    \item \textbf{Transformation of the basis}
    (\emph{\=a\'sraya-par\=av\d{r}tti}) predicted that the memory store
    requires periodic reorganization. \alaya{}'s \texttt{transform()}
    operation---deduplication, contradiction resolution, category
    discovery---implements this prediction.
    \item \textbf{Seed mutual conditioning} predicted that stored elements
    should activate related elements through weighted, use-adaptive links.
    \alaya{}'s Hebbian graph overlay with spreading activation implements
    this structural isomorphism directly.
\end{itemize}

The value of Yogac\=ara as an architectural guide is not mystical but
structural: it provides a coherent theory of memory-as-process that
identifies operations (perfuming, ripening, transformation) which neither
pure neuroscience models nor pure engineering intuition would prioritize
independently. The preference emergence mechanism, in particular, was
designed from the v\=asan\=a concept before we found empirical support in
the memory systems literature.

\subsection{Convergence with Neuroscience}

The parallels between Yogac\=ara and cognitive neuroscience are
especially striking because the two traditions developed independently,
separated by fifteen centuries and entirely different methodological
commitments.  Their convergence on the same structural patterns provides
independent triangulation for \alaya{}'s design principles.

\paragraph{Dual-system consolidation}
CLS theory~\cite{mcclelland1995} proposes fast-learning hippocampal
encoding with slow-learning neocortical integration through interleaved
replay.  The Yogac\=ara distinction between \=alaya-vij\~n\=ana (which
receives all impressions rapidly) and manas (which reflects
deliberatively) describes the same architectural split at the
phenomenological level.

\paragraph{Hebbian strengthening and weakening}
The \emph{Yogac\=arabh\=umi-\\'s\=astra} describes seeds as
strengthening through repeated activation (\cjk{增長}) and weakening
when not activated (\cjk{損減})---the same dynamics as LTP/LTD in
Hebb's framework~\cite{hebb1949, bliss1973}, expressed in
phenomenological rather than physiological vocabulary.

\paragraph{Spreading activation and seed mutual conditioning}
Collins and Loftus~\cite{collins1975} formalized spreading activation in
semantic networks.  The \emph{Yogac\=arabh\=umi-\\'s\=astra} describes
seeds as mutually conditioning (\cjk{互相薰習}): activation of one seed
creates conditions for the ripening of associated seeds.  The structural
isomorphism is precise: interconnected elements, use-dependent
facilitation, distance-based attenuation.

\paragraph{Competitive retrieval dynamics}
Anderson~\etal~\cite{anderson1994} demonstrated retrieval-induced
forgetting (RIF): retrieving one memory suppresses competitors.
Yogac\=ara describes an analogous dynamic among mutually opposed seeds
(\cjk{相違種子})~\cite{chengweishilun}.  Both predict that retrieval is
an active process that reshapes the accessibility landscape of memory.

\medskip
\noindent When three traditions separated by methodology, culture, and
fifteen centuries arrive at the same structural principles---dual-system
consolidation, use-dependent strengthening, associative spreading, and
competitive retrieval dynamics---those principles merit attention as
candidate universals of memory architecture.

\subsection{Graceful Degradation}

A deliberate design principle is that \alaya{} never fails due to missing
optional capabilities. The system operates in three degradation modes:

\begin{enumerate}[nosep]
    \item \textbf{Full capability:} Embeddings + LLM provider available.
    All retrieval signals (BM25, vector, graph), embedding-based clustering,
    LLM-generated category labels.
    \item \textbf{No embeddings:} BM25 + graph retrieval only. Category
    discovery uses graph-only clustering (shared Hebbian link overlap).
    \item \textbf{No LLM:} Consolidation is unavailable, but all other
    lifecycle operations function: forgetting, transformation (on existing
    semantic nodes), category discovery (from existing nodes), preference
    emergence (from accumulated impressions).
\end{enumerate}

This design reflects a practical reality: many deployment contexts
(offline agents, privacy-sensitive environments, cost-constrained settings)
cannot guarantee access to embedding models or LLMs for every memory
operation. The system should do the best it can with available resources
rather than failing entirely.

\subsection{Positioning in the Landscape}

The companion survey's capability scoring~\cite{hui2026taxonomy} evaluates
systems across eight discriminating features. \alaya{} implements all eight:
episodic memory, semantic memory, graph structure, forgetting,
consolidation, contradiction resolution, preference learning, and hybrid
retrieval fusion. Among the 88 surveyed systems, only 2.3\% score 8/8 on
this metric.

The distinctive characteristics relative to comparable systems are:
\begin{itemize}[nosep]
    \item vs.\ \textbf{Mem0}~\cite{choudhary2025mem0}: \alaya{} adds
    principled forgetting, graph-based retrieval, emergent (not extracted)
    preferences, and consolidation. Mem0 requires external services
    (Qdrant/Neo4j); \alaya{} requires none.
    \item vs.\ \textbf{Zep/Graphiti}~\cite{dempsey2025graphiti}: Both use
    knowledge graphs, but \alaya{}'s graph is Hebbian (reshapes through use)
    while Graphiti's is LLM-extracted (static after creation). \alaya{} adds
    forgetting and preference emergence.
    \item vs.\ \textbf{SYNAPSE}~\cite{jiang2026synapse}: Both implement
    CLS-inspired consolidation and activation-based retrieval. \alaya{} adds
    emergent ontology and the implicit (preference) store. SYNAPSE is
    evaluated on specific benchmarks; \alaya{} is a general-purpose library.
    \item vs.\ \textbf{Generative Agents}~\cite{park2023generative}: Both
    implement episodic-to-semantic consolidation (reflection). \alaya{} adds
    principled forgetting (Bjork vs.\ recency weighting), Hebbian graph
    evolution, and emergent categories.
\end{itemize}

\subsection{Limitations}

\alaya{} v0.1.0 has several limitations:
\begin{itemize}[nosep]
    \item \textbf{No end-to-end benchmark results:} We report baseline
    crossover analysis but have not yet evaluated \alaya{}'s full pipeline
    against LoCoMo/LongMemEval. This is ongoing work.
    \item \textbf{Flat categories only:} The emergent ontology discovers
    flat categories without hierarchy. Category hierarchy, prototype theory,
    and category seeds are planned for v0.2.0.
    \item \textbf{Single-agent only:} No multi-agent memory sharing,
    ownership semantics, or federated memory.
    \item \textbf{Thresholds are hand-tuned:} Cosine similarity thresholds
    for category assignment (0.6), discovery (0.7), and merge (0.85) are
    based on reasonable defaults, not learned from data.
    \item \textbf{SQLite scalability:} While SQLite handles millions of
    records efficiently, the in-process architecture limits concurrent write
    access to a single process.
\end{itemize}
